<!-- ========================================= -->
<!-- Chapter: 35-matching-methods.Rmd -->
<!-- ========================================= -->


# 매칭 방법 {#sec-matching-methods}

매칭은 관측된 특성이 유사한 처치집단과 비교집단을 구성하여 잠재적인 편향의 원인을 제거하거나 최소화하는 전략이다. 이를 통해 두 집단의 결과에서 관측되는 차이를 다른 요인보다 처치 자체에 더 신뢰성 있게 귀속할 수 있다. 관찰연구에서는 특히 여러 처치 전 결과를 이용할 수 있을 때 선택 편향 문제를 다루기 위해 매칭을 [Difference-in-Differences] (DiD) 기법과 결합하는 경우가 많다.

> 매칭은 "처치집단과 통제집단의 공변량 분포를 같게(또는 '균형을') 만들려는 모든 방법"으로 정의된다. [@stuart2010matching, pp. 1]

매칭은 다음과 같은 경우에 특히 유용하다.

-   후속 연구처럼 결과가 **아직 관측되지 않았고**, 균형 잡힌 처치/통제집단을 구성하려는 경우.
-   결과를 **이용할 수 있지만**, **모형 의존성을 줄이고** **강건성을 높이려는** 경우.

> 개념적으로 매칭은 **결측자료(missing data)**의 관점에서도 이해할 수 있다. 어떤 단위에 대해서도 두 잠재적 결과 $(Y_i^T, Y_i^C)$를 동시에 관측할 수 없기 때문이다. 따라서 이 주제는 [Imputation (Missing Data)]와 밀접하게 관련된다.

## 서론과 동기

### 왜 매칭하는가?

많은 관찰연구에서 연구자는 무작위 배정이라는 이점을 누릴 수 없다. 연구대상(사람, 기업, 학교 등)은 일반적으로 특정 관측·비관측 특성에 따라 처치를 선택하거나 처치 대상으로 선택된다. 이 과정에서 인과추론을 교란하는 체계적 차이(선택 편향)가 발생할 수 있다. 매칭은 처치 단위와 비처치(통제) 단위 사이에서 이러한 관측 특성을 "균형화"하여 무작위 실험을 근사하려는 방법이다.

-   **목표:** 처치 대상과 통제 대상의 공변량이 충분히 비교 가능하도록 하여 모형 의존성을 줄이고 인과효과를 명확히 한다.
-   **과제:** 매칭으로 관측 공변량의 균형을 달성하더라도 **비관측** 교란요인은 여전히 식별을 위협한다(즉, 매칭은 selection on observables 식별 전략일 뿐이다). 매칭이 비관측 변수로 인한 편향을 자동으로 해결해 주는 것은 아니다.

관찰연구에서 인과추론이 어려운 이유를 이해하려면 다음을 살펴보자.

$$ 
\begin{aligned} E(Y_i^T | T) - E(Y_i^C | C) &= E(Y_i^T - Y_i^C | T) + \underbrace{[E(Y_i^C | T) - E(Y_i^C | C)]}_{\text{Selection Bias}} \\ \end{aligned} $$

-   $E(Y_i^T - Y_i^C | T)$ 항은 **인과효과**(구체적으로 ATT)이다.
-   $E(Y_i^C | T) - E(Y_i^C | C)$ 항은 처치집단과 통제집단의 처치받지 않은 잠재적 결과에 존재하는 체계적 차이로 인한 **선택 편향**을 나타낸다.

무작위 배정은 다음을 보장한다.

$$ E(Y_i^C | T) = E(Y_i^C | C) $$

이는 선택 편향을 제거한다. 그러나 관찰자료에서는 이 등식이 성립하는 경우가 드물다.

매칭은 공변량 $X$에 **조건부화(conditioning)**하여 무작위 배정을 모방하는 것을 목표로 한다.

$$ E(Y_i^C | X, T) = E(Y_i^C | X, C) $$

예를 들어 성향점수 매칭은 **성향점수(propensity score)** $P(X)$에 조건부화하여 이러한 균형을 달성한다.

$$ E(Y_i^C | P(X), T) = E(Y_i^C | P(X), C) $$

자세한 내용은 [Propensity Scores](#sec-propensity-scores)를 참조하라.

매칭에서 [Average Treatment Effect] (ATE)는 일반적으로 다음과 같이 추정한다.

$$ \frac{1}{N_T} \sum_{i=1}^{N_T} \left(Y_i^T - \frac{1}{N_{C_i}} \sum_{j \in \mathcal{C}_i} Y_j^C\right) $$

여기서 $\mathcal{C}_i$는 처치 단위 $i$에 매칭된 통제 단위를 나타낸다.

매칭의 표준오차

-   매칭에는 ATE나 ATT의 표준오차에 대한 폐쇄형 해가 없다.
-   따라서 불확실성은 부트스트래핑으로 추정한다.

> **참고**: 매칭은 매칭되지 않은 관측치를 버려 유효 표본 크기를 줄이므로 [OLS][Ordinary Least Squares]보다 표준오차가 커지는 경향이 있다.

------------------------------------------------------------------------

### "가지치기"로서의 매칭

매칭은 "**가지치기(pruning)**"(**전처리(preprocessing)** 단계)로 볼 수 있다 [@king2017balance]. 분석에 앞서 매칭되지 않았거나 매칭 품질이 낮은 단위를 **가지쳐 내어**, 모형 의존성을 줄이는 것이 목표다.

매칭 없음:

-   불균형 자료 → 모형 의존성 → 연구자의 재량 → 편향된 추정치

매칭 적용:

-   균형 자료 → 재량 감소 → 더 신뢰할 수 있는 인과추론

+-----------------------+------------------------+----------------------+
| 공변량 균형 | 완전 무작위화 | 완전 정확 매칭 |
+=======================+========================+======================+
| **관측**          | 평균적으로             | 정확                |
+-----------------------+------------------------+----------------------+
| **비관측**        | 평균적으로             | 평균적으로           |
+-----------------------+------------------------+----------------------+

: 설계별 균형 정도

완전 블록화 또는 정확 매칭 설계는 이러한 기준에서 완전 무작위화와 *동등할 뿐 아니라* 이를 능가한다. 공변량을 우연에 맡기지 않고 일치시키므로 불균형이 줄어들고, 분석자가 회귀로 기초선 차이를 임시로 덮을 필요가 없어 모형 의존성이 낮아진다. 잔차 변동이 집단 차이에 잠식되지 않으므로 효율성과 통계적 검정력이 높아지고, 관측 가능한 이질성으로 인한 편향은 평균화가 아니라 설계에 의해 제거된다. 그 결과 추정치는 모형 설정 선택에 더 강건해지며, 초기 설계 비용은 더 깔끔한 추론과 더 작은 필요 표본으로 대체로 회수된다.

------------------------------------------------------------------------

### DiD와 매칭

여러 처치 전 기간을 이용할 수 있다면 매칭을 DiD와 효과적으로 결합할 수 있다. 이러한 설계는 다음 가정하에서 선택 편향을 보정하는 데 도움이 된다.

-   선택 편향이 처치일을 중심으로 대칭이면 표준 DID(처치일을 중심으로 대칭적으로 구현)는 여전히 일관적이다 [@chabe2015analysis].
-   선택 편향이 비대칭이면 @chabe2015analysis의 시뮬레이션은 대칭 DID가 매칭 단독보다 여전히 우수함을 보여준다. 다만 처치 전 관측치가 많을수록 매칭 성능이 향상될 수 있다.

요컨대 매칭은 만능 해법은 아니지만, DiD나 다른 인과 추정 방법을 적용하기 전에 유용한 전처리 단계가 되는 경우가 많다 [@smith2005does].

------------------------------------------------------------------------

## 핵심 가정

매칭은 [selection on observables](#sec-selection-on-observables)의 근간을 이루는 표준 가정, 즉 back-door criterion( [Assumptions for Identifying Treatment Effects] 참조)에 의존한다. 이 가정이 성립하면 관측 공변량이 비교 가능한 처치집단과 통제집단을 구성하여 유효한 인과효과 추정치를 얻을 수 있다.

1.  **Strong** [Conditional Ignorability Assumption] (Unconfoundedness)

이는 **숨은 편향 없음(no hidden bias)** 또는 **무시 가능성(ignorability)** 가정이라고도 한다.

$$
(Y(0), Y(1)) \perp T \big| X
$$

이는 공변량 $X$에 조건부일 때 처치 배정이 잠재적 결과와 독립임을 뜻한다. 즉, $X$를 보정하면 비관측 교란요인이 없다는 의미다.

-   이 가정은 검정할 수 없지만, 관련된 모든 교란요인이 관측되어 $X$에 포함되어 있을 때 더 타당하다.
-   비관측 공변량이 관측 공변량과 높은 상관을 가지면 대체로 근사적으로 성립한다.
-   비관측 변수가 $X$와 무관하다면 다음을 수행할 수 있다.
    -   추정치의 강건성을 검토하는 민감도 분석을 실시한다.
    -   설계 민감도 기법을 적용한다. 비관측 교란이 의심되면 [@heller2009split]의 설계 민감도 접근법이나 경계 접근법(예: `rbounds` R 패키지)으로 숨은 편향에 대한 결과의 강건성을 검토할 수 있다.

> 이는 매칭의 초석이 되는 가정이다. 이 가정이 없으면 관찰자료에서의 인과추론은 일반적으로 타당하지 않다.

------------------------------------------------------------------------

2.  [Overlap (Positivity) Assumption] **(Common Support)**

$$
0 < P(T=1 \mid X) < 1 \quad \forall X
$$

이 조건은 공변량 $X$의 모든 값에서 처치와 통제를 받을 확률이 모두 양수임을 보장한다.

-   이 가정이 실패하면 공변량 공간의 일부 영역에 처치 단위나 통제 단위가 없어 비교가 불가능해진다.
-   매칭은 공통 지지 영역 밖의 단위를 버려 이 가정을 강제한다.

> 이 가지치기 단계는 매칭의 장점이자 한계다. 일반화 가능성을 희생하는 대신 내적 타당도를 높이기 때문이다.

------------------------------------------------------------------------

3.  [Stable Unit Treatment Value Assumption] (SUTVA)

SUTVA는 다음을 요구한다.

-   어떤 개별 단위의 잠재적 결과도 다른 단위의 처치 배정에 의존하지 않는다.

즉, 단위 사이에 **간섭(interference)이나 파급효과(spillover effect)**가 없다.

-   Mathematically, $Y_i(T_i)$ depends only on $T_i$, not on $T_j$ for any $j \neq i$.
-   다음과 같은 상황에서는 위반이 발생할 수 있다.
    -   교육(동료 효과)
    -   역학(질병 전파)
    -   마케팅(네트워크 영향)

> 파급효과가 알려진 경우에는 상호작용을 줄이거나 간섭을 명시적으로 모형화해야 한다.

------------------------------------------------------------------------

매칭 가정 요약

+------------------------------+-------------------------------------------------------------------------------------+----------------------------------------------+
| 가정                   | 설명                                                                         | 표기                                     |
+==============================+=====================================================================================+==============================================+
| **조건부 무시 가능성** | 공변량 조건부화 후 숨은 교란 없음                              | $(Y(0), Y(1)) \perp T \mid X$                |
+------------------------------+-------------------------------------------------------------------------------------+----------------------------------------------+
| **중첩(양성성)**     | 각 단위가 처치와 통제에 배정될 확률이 0이 아님            | $0 < P(T=1 \mid X) < 1$                      |
+------------------------------+-------------------------------------------------------------------------------------+----------------------------------------------+
| **SUTVA**                    | 단위 간 간섭 없음; 한 단위의 결과가 다른 단위의 처치에 영향받지 않음 | $Y_i(T_i)$ unaffected by $T_j$ for $j \ne i$ |
+------------------------------+-------------------------------------------------------------------------------------+----------------------------------------------+

이 세 가정은 매칭 방법을 이용한 유효한 인과추론의 토대다.

------------------------------------------------------------------------

## 일반화 프레임워크

가정을 정했으면 매칭 추정대상을 정확히 쓰고 특정 절차가 어떤 평균효과(ATE 또는 ATT)를 회복하는지 논의할 수 있도록 표기를 정한다.

다음과 같이 두자.

-   $P_t$, $P_c$: 처치 모집단과 통제 모집단
-   $N_t$, $N_c$: $P_t$, $P_c$에서 추출한 무작위 표본
-   $\mu_i$, $\Sigma_i$: 집단 $i \in \{t, c\}$의 $p$개 공변량에 대한 평균과 공분산 행렬
-   $X_j$: 개인 $j$의 공변량 벡터
-   $T_j \in \{0, 1\}$: 처치 지시변수(1 = 처치, 0 = 통제)
-   $Y_j$: 관측 결과
-   $N_t < N_c$라고 가정한다(즉, 처치 단위보다 통제 단위가 많다).

조건부 처치효과는 다음과 같다.

$$
\tau(x) = R_1(x) - R_0(x), \quad \text{where } R_1(x) = E[Y(1) \mid X = x], \quad R_0(x) = E[Y(0) \mid X = x]
$$

**동질적 처치효과**를 가정하면 모든 $x$에 대해 $\tau(x) = \tau$다. 이는 처치효과에 대한 동질성 가정이며, 반사실적 *시간 추세*에 대한 별도의 제약인 [DiD](#sec-difference-in-differences)의 평행 추세 가정과는 다르다. 동질성 가정을 완화해도 $X$의 분포에 대한 **평균** 효과는 추정할 수 있다.

### 일반적인 추정대상

-   [Average Treatment Effect] (ATE): 모든 단위에 대한 평균 인과효과.
-   [Average Treatment Effect on the Treated] (ATT): 처치 단위에 한정한 인과효과.

------------------------------------------------------------------------

## 매칭 단계

매칭 분석은 네 단계의 파이프라인으로 구현한다. 두 단위의 근접성을 측정할 방법을 고르고, 짝을 구성할 알고리즘을 선택하고, 그 결과로 얻은 비교집단이 실제로 균형을 이루는지 확인한 다음, 처치효과를 추정한다. 아래의 세 요소인 성향점수, 거리 측도, 그리고 무시 가능성을 만족한다고 가정한 공변량 집합은 이후 단계마다 반복해서 사용된다.

-   성향점수: $P(T=1|X)$를 요약한다.
-   거리 측도: 유사성을 측정한다.
-   공변량: 무시 가능성을 만족한다고 가정한다.

### 1단계: "근접성" 정의하기(거리 측도)

매칭에는 처치 단위와 통제 단위의 유사성을 정의할 거리 측도가 필요하다.

#### 변수 선택 지침

-   조건부 무시 가능성을 뒷받침하도록 가능한 한 많은 처치 전 공변량을 포함한다.
-   편향을 유발하는 처치 후 변수를 피한다.
-   처치와는 상관이 없지만 결과와 높은 상관을 갖는 변수(예: 매개변수)에 주의한다.
-   처치와 결과 모두와 상관이 없는 변수라면 포함 비용은 작다.

#### 거리 측도

+----------------------------+----------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------+
| 방법                     | 공식                                            | 비고                                                                                                                                    |
+============================+====================================================+==========================================================================================================================================+
| **정확 매칭**         | $D_{ij} = 0$ if $X_i = X_j$, else $\infty$         | 저차원에서만 가능하며 [Coarsened Exact Matching]으로 완화할 수 있음                                                           |
+----------------------------+----------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------+
| **마할라노비스 거리**   | $D_{ij} = (X_i - X_j)' \Sigma^{-1}(X_i - X_j)$     | $\Sigma$는 ATT가 관심 대상이면 통제집단에서, ATE가 관심 대상이면 통합 표본에서 구한 $X$의 분산-공분산 행렬이며 차원에 민감함 |
+----------------------------+----------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------+
| **성향점수**       | $D_{ij} = |e_i - e_j|$                             | $e_k$는 단위 $k$의 추정 성향점수 $P(T=1 \mid X_k)$다.                                                            |
|                            |                                                    |                                                                                                                                          |
|                            |                                                    | 고급 방법: **예후점수(Prognostic scores)** [@hansen2008prognostic]는 $E[Y(0)|X]$를 모형화해야 하므로 결과 모형에 의존한다.               |
+----------------------------+----------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------+
| **로짓 성향점수** | $D_{ij} = |\text{logit}(e_i) - \text{logit}(e_j)|$ | 분포의 꼬리에서 더 안정적                                                                                                     |
+----------------------------+----------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------+

> 팁: 고차원에서는 정확 매칭과 마할라노비스 매칭의 성능이 낮다. 마할라노비스 거리와 성향점수 캘리퍼를 결합하면 강건성을 높일 수 있다 [@rubin2000combining].

종단 자료를 위한 고급 방법:

-   주변 구조 모형(Marginal Structural Models): 시간가변 처치를 위한 방법 [@robins2000marginal]
-   균형 위험집합 매칭(Balanced Risk Set Matching): 생존분석을 위한 방법 [@li2001balanced]

------------------------------------------------------------------------

### 2단계: 매칭 알고리즘

매칭 알고리즘의 선택에는 익숙한 절충이 따른다. 모든 공변량에서 거의 정확한 일치를 요구하는 공격적인 알고리즘(예: 정확 매칭이나 세밀한 [Coarsened Exact Matching](#sec-cem))은 높은 균형을 제공하지만 표본 대부분을 가지쳐 내어 좁은 모집단 일부에만 적용되는 추정대상을 남기고 분산을 키울 수 있다. 완화된 알고리즘(예: 단일 성향점수에 대한 최근접 이웃 매칭)은 더 많은 단위를 유지하고 표준오차가 더 작은 추정치를 제공하지만 잔여 불균형이 편향으로 이어질 수 있다. 각 알고리즘이 균형과 표본 유지 사이의 경계에서 한 지점을 선택한다고 생각해야 한다. 뒤에서 설명할 [`MatchingFrontier`](#sec-matching-methods) 도구는 이 경계를 명시적으로 보여준다. 아래 알고리즘은 주로 이 경계를 탐색하는 방식에서 차이가 난다.

1.  Nearest Neighbor Matching

최근접 이웃 매칭은 가장 일반적인 기본 방법이다. 각 처치 단위를 선택한 거리 측도에서 이용 가능한 가장 가까운 통제 단위와 짝지어, 쌍 무작위 실험을 모방하는 직관적인 매칭 표본을 만든다. 구현과 분석은 쉽지만 품질은 거리 측도와 통제 단위의 공급량에 전적으로 좌우된다. 통제가 부족하거나 공변량이 많으면 "가장 가까운" 단위도 실제로는 멀 수 있어 편향-분산 절충이 더욱 중요해진다.

-   탐욕적 매칭: 빠르지만 통제 단위 경쟁이 있으면 최적이 아니다.
-   최적 매칭: 모든 쌍의 전역 거리를 최소화한다.
-   비율 매칭(k:1): 통제가 처치보다 많을 때 유용하며 편향과 분산의 절충을 고려해 $k$를 선택한다 [@rubin1996matching].
-   With vs. without replacement:
    -   복원 추출: 매칭 품질을 높이지만 분석에 빈도 가중치가 필요하다.
    -   비복원 추출: 더 단순하지만 유연성이 낮다.

2.  Subclassification, Full Matching, and Weighting

이 방법들은 분수 가중치를 부여하여 최근접 이웃 접근법을 일반화한다. 짝을 찾지 못한 통제 단위를 버리는 대신 모든 단위에 다른 집단의 단위와의 유사성을 반영하는 가중치를 부여한다. 그 결과 더 완화된 형태의 가지치기가 이루어져 전체 표본을 유지하는 경우가 많으며, "매칭 쌍"의 해석 가능성을 일부 포기하고 효율성을 얻는다. 처치집단과 통제집단의 정보를 대칭적으로 사용하므로 목표 추정대상이 [ATT](#sec-counterfactual-estimators)보다 [ATE](#sec-counterfactual-estimators)일 때 가중 추정량이 특히 매력적이다.

-   하위분류: 성향점수에 따라 층(예: 5분위)으로 나눈다.
-   완전 매칭: 각 처치 단위를 가중 통제집단(반대 방향도 동일)과 매칭하여 집합 내 평균 거리를 최소화한다.
-   가중: 가중 기법은 성향점수로 ATE를 추정한다. 다만 가중치가 극단적이면 분산이 커질 수 있다. 이는 기저 확률이 아니라 추정 절차 자체에서 비롯된다. 이를 해결하려면 성향점수를 가중이나 매칭에 사용할 때 (1) 가중치 절단 또는 (2) 이중 강건 방법을 사용할 수 있다.
    -   Inverse Probability of Treatment Weighting (IPTW): $$
        w_i = \frac{T_i}{\hat{e}_i} + \frac{1 - T_i}{1 - \hat{e}_i}
        $$
    -   Odds weighting: $$
        w_i = T_i + (1 - T_i)\frac{\hat{e}_i}{1 - \hat{e}_i}
        $$
    -   커널 가중: 통제집단에 대한 평활 평균(경제학에서 널리 사용).
    -   절단 및 이중 강건 방법: 극단적 가중치로 인한 분산을 줄인다.

3.  Assessing Common Support

-   성향점수 히스토그램으로 중첩을 시각화한다.
-   $X$의 볼록 껍질 밖에 있는 단위(즉, 매칭되지 않는 영역)는 버릴 수 있다.
-   중첩이 부족하면 일부 비교가 경험적 매칭이 아니라 외삽임을 의미한다.

------------------------------------------------------------------------

### 3단계: 매칭 품질 진단

#### 균형 진단

매칭은 처치 단위와 통제 단위의 공변량 분포를 균형화하는 것을 목표로 한다. 매칭이 잘 된 표본은 다음을 만족한다.

$$
\tilde{p}(X \mid T=1) \approx \tilde{p}(X \mid T=0)
$$

여기서 $\tilde{p}$는 경험적 분포다.

1.  수치 점검

-   평균의 표준화 차이(가장 일반적): $< 0.1$이어야 한다.
-   성향점수의 표준화 차이: $< 0.25$이어야 한다 [@rubin2001using].
-   성향점수의 분산비: 0.5와 2.0 사이 [@rubin2001using].
-   각 공변량을 성향점수에 회귀한 후 잔차의 분산(처치 대 통제).

> p값을 진단으로 사용하지 말라. p값은 균형과 통계적 검정력을 혼동하며 표본 크기에 민감하다.

2.  그래픽 진단

-   경험적 분포 도표
-   분위수-분위수(QQ) 도표
-   러브 도표: 매칭 전후 표준화 차이를 요약한다.

------------------------------------------------------------------------

### 4단계: 처치효과 추정

1.  매칭 후

-   복원 추출을 사용하는 k:1 매칭에서는 통제 단위의 재사용을 조정하기 위해 가중치를 사용한다.
-   매칭 표본에 회귀 보정을 적용하여 정밀도를 높이고 잔여 불균형을 조정한다.

2.  하위분류 또는 완전 매칭 후

-   ATT: 하위분류별 추정치를 처치 단위 수로 가중한다.
-   ATE: 하위분류별 전체 단위 수로 가중한다.

3.  분산 추정

다음 두 요소의 불확실성을 모두 반영해야 한다.

1.  매칭 절차(표본 추출 및 거리 계산)(3단계)
2.  결과 모형(회귀, 평균 차이 등)(4단계)

대개 부트스트래핑으로 추정한다.

------------------------------------------------------------------------

## 특별 고려사항

1.  결측자료 처리

-   일반화 부스팅 모형이나 다중대치를 사용한다 [@qu2009propensity].

2.  무시 가능성 위반

비관측 요인이 처치에 편향을 유발할 때의 전략:

-   처치 전 결과 측정치를 사용한다.
-   여러 통제집단을 비교한다.
-   민감도 분석을 수행한다.
    -   관측된 효과를 0으로 만들기 위해 비관측 교란요인과 처치 및 결과 사이의 상관을 정량화한다.

------------------------------------------------------------------------

## 매칭 전략 선택

모든 응용에서 우월한 단일 매칭 알고리즘은 없다. 적절한 선택은 목표 추정대상( [counterfactual estimators](#sec-counterfactual-estimators) 참조), 처치 단위와 통제 단위의 비율, 공변량의 차원과 유형, 그리고 균형과 표본 유지 사이의 절충에서 분석자가 어느 지점을 택하는지에 달려 있다. 아래 지침은 합리적인 기본값을 제시하며, 선택을 검증하는 [balance checks](#sec-balance-checks)와 함께 읽어야 한다.

### 추정대상에 따른 선택

-   ATE: IPTW 또는 완전 매칭을 사용한다.
-   ATT:
    -   통제가 많은 경우($N_c > 3N_t$): 비복원 k:1 최근접 이웃 매칭.
    -   통제가 적은 경우: 하위분류, 완전 매칭 또는 오즈 가중.

### 진단 결과에 따른 선택

-   균형이면 매칭 표본에 회귀를 적용한다.
-   소수 공변량에서만 불균형이면 해당 공변량에 마할라노비스 매칭을 적용한다.
-   다수 공변량에서 불균형이면 복원 k:1 매칭을 시도한다.

### 선택 기준

-   여러 공변량의 표준화 차이를 최소화한다.
-   특히 예후 공변량을 우선한다.
-   불균형이 큰($>0.25$) 공변량의 수를 최소화한다 [@diamond2013genetic].

매칭에 모든 경우에 맞는 단일 방법은 없다. 목표 추정대상, 자료 구조, 진단 결과를 바탕으로 방법을 선택한다.

A short translation of the algorithmic choices into applied guidance: reach for [propensity score](#sec-propensity-scores) methods when covariates are many but the propensity model is plausibly well specified and the target is a population estimand; reach for [Mahalanobis distance](#sec-mahalanobis) when covariates are few, continuous, and well measured, and when emulating a fully blocked design is the priority; reach for [Coarsened Exact Matching](#sec-cem) when the analyst wants transparent, ex ante control over imbalance and is willing to coarsen continuous covariates into substantively meaningful bins; reach for [Genetic Matching](#sec-genetic-matching) when distance-metric specification is itself uncertain and the dataset is large enough to absorb its computational cost; reach for entropy balancing or weighting when treated units are few and discarding any of them would be unacceptable. Sensitivity to unobserved confounding (covered in [Rosenbaum bounds](#sec-rosenbaum-bounds) and the broader discussion of [selection on unobservables](#sec-selection-on-unobservables)) should accompany any matching analysis, since none of these algorithms can rescue a violation of [conditional ignorability](#sec-conditional-ignorability-assumption).

------------------------------------------------------------------------

## 역확률 가중치 {#sec-inverse-probability-weighting}

Weighting deserves a closer look than the brief mention it received among the matching algorithms above, because it is the direct bridge between matching and the estimating equations used by modern doubly robust and machine-learning estimators. The central insight is that inverse probability weighting (IPW) is not a causal invention at all. It is the Horvitz-Thompson estimator of survey sampling [@horvitz1952generalization], applied to the missing data problem that treatment assignment creates.

### Horvitz-Thompson의 기초

In sampling theory, a finite population of units $i = 1, \dots, N$ carries fixed attributes $Y_i$, and each unit is sampled with known inclusion probability $\pi_i = P(S_i = 1)$. The Horvitz-Thompson estimator of the population mean is

$$
\hat{\mu}_{HT} = \frac{1}{N} \sum_{i=1}^N \frac{S_i Y_i}{\pi_i},
$$

which is unbiased because $E[S_i Y_i / \pi_i] = Y_i$. Each observed unit stands in for $1/\pi_i$ population units, so rarely sampled units are upweighted to undo the selection.

Causal inference under the potential outcomes framework poses the same problem. Each unit has two fixed potential outcomes $Y_i(1)$ and $Y_i(0)$, but treatment assignment reveals only one of them. Treatment plays the role of the sampling indicator, and the propensity score $e(X_i) = P(D_i = 1 \mid X_i)$ plays the role of the inclusion probability. Because $E[D / e(X) \mid X] = 1$, we obtain

$$
E[Y(1)] = E\left[\frac{D\, Y}{e(X)}\right], \qquad
E[Y(0)] = E\left[\frac{(1-D)\, Y}{1 - e(X)}\right],
$$

and the sample analog

$$
\widehat{ATE}_{IPW} = \frac{1}{N} \sum_{i=1}^N \left( \frac{D_i Y_i}{\hat{e}(X_i)} - \frac{(1 - D_i) Y_i}{1 - \hat{e}(X_i)} \right)
$$

is algebraically a Horvitz-Thompson estimator applied separately to the treated and control potential-outcome populations. The correspondence is exact:

| Sampling theory | Causal inference |
|---------------------------------|---------------------------------------|
| Population units | Units with potential outcomes |
| Sampling indicator $S_i$ | Treatment indicator $D_i$ |
| Inclusion probability $\pi_i$ | Propensity score $e(X_i)$ |
| Missing by sampling design | Missing counterfactual |
| Positive inclusion probability | Overlap, $0 < e(X) < 1$ |
| Selection on known design | [Selection on observables](#sec-selection-on-observables) |

This mapping clarifies two things at once. First, the identification assumptions of IPW are sampling assumptions in disguise: unconfoundedness is the analog of a known selection mechanism, and overlap is the analog of every unit having a positive chance of being sampled. Horvitz-Thompson fails when $\pi_i = 0$, and IPW fails when $e(X)$ hits 0 or 1. Second, the variance pathologies of IPW are structural rather than incidental. The Horvitz-Thompson estimator is unbiased but can have enormous variance when some inclusion probabilities are tiny, and the same heavy tails in $1/e(X)$ dominate finite samples in causal applications. Trimming and stabilized weights are variance-control devices, not identification tools.

### 가중치, 추정대상, 밀도비 관점

Different target estimands correspond to different weight functions, and the cleanest way to see this is through density ratios. Let $f(x)$ denote the marginal covariate density and $f(x \mid D=1)$, $f(x \mid D=0)$ the group-specific densities. Bayes' rule gives $f(x) = P(D=1)\, f(x \mid D=1) / e(x)$, so reweighting treated units by $1/e(x)$ transforms their covariate distribution into that of the full population. The same logic for other targets yields the standard menu:

| Target estimand | Treated weight | Control weight | Main risk |
|------------------|------------------|------------------|-------------------|
| ATE | $1/e(X)$ | $1/(1-e(X))$ | Extreme weights on both sides |
| ATT | $1$ | $e(X)/(1-e(X))$ | Extreme weights on controls |
| ATO (overlap) | $1-e(X)$ | $e(X)$ | None, but estimand changes |

A useful summary of how much information survives reweighting is the effective sample size,

$$
ESS = \frac{\left(\sum_i w_i\right)^2}{\sum_i w_i^2},
$$

which shrinks as the weights become more dispersed. Even a nominally large control group contributes little if the target population forces a handful of controls with large $e(X)$ to carry most of the weight.

### 중첩 가중치

When overlap is limited, plain ATE weights explode near $e(X) = 0$ or $1$. The overlap weights of @li2018balancing replace $1/e(X)$ and $1/(1-e(X))$ with $1-e(X)$ for treated units and $e(X)$ for controls. Three properties make them attractive in practice. They are bounded by construction, so no observation can dominate. They achieve exact mean balance on any covariate included in a logistic propensity model, a small-sample property rather than an asymptotic one. And they minimize the asymptotic variance of the weighted estimator within the class of balancing weights. The price is a change of estimand: overlap weights target the average treatment effect in the overlap population, the units most in equipoise between treatment and control, rather than the full-population ATE. In settings such as clinical decision-making or marginal policy expansion this subpopulation is often the one of substantive interest, which is why overlap weighting has moved quickly from a variance fix to a first-choice design.

### 간단한 시뮬레이션

The simulation below generates a confounded treatment with true ATE equal to 1 and compares the naive difference in means, regression adjustment, ATE-targeted IPW, and overlap weighting, along with the effective sample size each weighting scheme retains in the control group.

```{r ipw-overlap-simulation}
set.seed(123)
n  <- 5000
x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- rbinom(n, 1, 0.4)

# Confounded assignment and outcomes, true ATE = 1
e_true <- plogis(-0.5 + 0.8 * x1 - 0.6 * x2 + 0.4 * x3)
d  <- rbinom(n, 1, e_true)
y0 <- 2 + 1.0 * x1 + 0.5 * x2 + 0.3 * x3 + rnorm(n)
y  <- ifelse(d == 1, y0 + 1, y0)
dat <- data.frame(y, d, x1, x2, x3)

ps_fit <- glm(d ~ x1 + x2 + x3, data = dat, family = binomial())
e_hat  <- fitted(ps_fit)

ate_naive <- mean(dat$y[dat$d == 1]) - mean(dat$y[dat$d == 0])
ate_reg   <- coef(lm(y ~ d + x1 + x2 + x3, data = dat))["d"]

# ATE weights (Horvitz-Thompson)
w_ipw <- ifelse(dat$d == 1, 1 / e_hat, 1 / (1 - e_hat))
ate_ipw <- weighted.mean(dat$y[dat$d == 1], w_ipw[dat$d == 1]) -
    weighted.mean(dat$y[dat$d == 0], w_ipw[dat$d == 0])

# Overlap weights
w_ow <- ifelse(dat$d == 1, 1 - e_hat, e_hat)
ate_ow <- weighted.mean(dat$y[dat$d == 1], w_ow[dat$d == 1]) -
    weighted.mean(dat$y[dat$d == 0], w_ow[dat$d == 0])

ess <- function(w) sum(w)^2 / sum(w^2)

data.frame(
    estimator = c("Naive", "Regression", "IPW (ATE)", "Overlap"),
    estimate  = round(c(ate_naive, ate_reg, ate_ipw, ate_ow), 3),
    ess_controls = round(c(NA, NA, ess(w_ipw[dat$d == 0]),
                           ess(w_ow[dat$d == 0])))
)
```

The naive comparison overstates the effect because treatment is concentrated where baseline outcomes are high, while regression adjustment, IPW, and overlap weighting all recover the truth. In this design overlap is healthy, so IPW and overlap weights perform similarly; as overlap deteriorates, the IPW estimate becomes erratic while the overlap-weighted estimate remains stable, at the cost of answering the question for the overlap population only. In applied work the `WeightIt` package estimates all of these weighting schemes with balance diagnostics, and weighting is usually paired with an outcome model in a doubly robust estimator, a combination developed in [Matching vs. Regression] below and in the machine-learning treatment of Chapter \@ref(sec-ml-causal).

------------------------------------------------------------------------

## 매칭과 회귀의 비교

매칭과 회귀는 관찰연구에서 공변량 $X$의 차이를 보정하고 인과효과를 추정하는 두 핵심 전략이다. 두 방법 모두 교란으로 인한 편향 제거를 목표로 하지만, 관측치에 가중치를 부여하는 방식, 함수형 가정을 다루는 방식, 공변량 균형을 처리하는 방식에서 서로 다르게 접근한다.

어느 방법도 비관측 교란 문제를 해결할 수는 없지만, 주의 깊게 사용하고 적절한 진단으로 뒷받침하면 각각 강력한 도구가 될 수 있다.

-   매칭은 비교 가능한 단위만 남기도록 자료를 가지쳐 공변량 균형을 강조한다. 비모수적이며 ATT에 초점을 둔다.
-   회귀(일반적으로 [OLS][Ordinary Least Squares])는 함수형을 강조하고 모형 기반 보정을 허용하므로 ATE와 처치의 연속적 효과 또는 상호작용 효과를 추정할 수 있다.

매칭과 회귀는 추정 과정에서 관측치에 암묵적 또는 명시적 가중치를 부여한다.

-   매칭: 처치 단위가 많은 층의 관측치에 더 큰 가중치를 부여하여 ATT 추정대상에 부합한다.
-   [OLS][Ordinary Least Squares] 회귀: 처치 배정의 분산이 가장 큰 층, 즉 처치와 통제가 대략 균형을 이루는 층(50/50에 가까운 경우)에 더 큰 가중치를 둔다.

그 결과 추정대상과 민감도가 달라진다.

> 중요 주의사항: [OLS][Ordinary Least Squares] 추정치가 비관측 교란으로 편향되어 있다면 매칭 추정치도 편향될 가능성이 높다. 두 방법 모두 [selection on observables](#sec-selection-on-observables) 가정에 의존한다.

특히 ATT 추정에서 매칭과 회귀의 추정대상이 어떻게 다른지 살펴본다.

### 매칭 추정대상

처치집단에 대한 처치효과를 구한다고 하자.

$$
\delta_{\text{TOT}} = E[Y_{1i} - Y_{0i} \mid D_i = 1]
$$

[Law of Iterated Expectation]을 사용하면:

$$
\delta_{\text{TOT}} = E\left[ E[Y_{1i} \mid X_i, D_i = 1] - E[Y_{0i} \mid X_i, D_i = 1] \mid D_i = 1 \right]
$$

조건부 독립을 가정하면:

$$
E[Y_{0i} \mid X_i, D_i = 0] = E[Y_{0i} \mid X_i, D_i = 1]
$$

그러면,

$$
\begin{aligned}
\delta_{TOT} &= E [ E[ Y_{1i} | X_i, D_i = 1] - E[ Y_{0i}|X_i, D_i = 0 ]|D_i = 1 ] \\
&= E\left[ E[Y_i \mid X_i, D_i = 1] - E[Y_i \mid X_i, D_i = 0] \mid D_i = 1 \right] \\
&= E[\delta_X |D_i = 1]
\end{aligned}
$$

$\delta_X$는 공변량 값 $X_i$에서의 $X$별 평균 차이다.

$X_i$가 이산형이면 매칭 추정대상은 다음과 같다.

$$
\delta_M = \sum_x \delta_x P(X_i = x \mid D_i = 1)
$$

여기서 $P(X_i = x |D_i = 1)$은 $D_i = 1$일 때 $X_i$의 확률질량함수다.

베이즈 규칙에 따르면:

$$
P(X_i = x \mid D_i = 1) = \frac{P(D_i = 1 \mid X_i = x) P(X_i = x)}{P(D_i = 1)}
$$

So,

$$
\begin{aligned}
\delta_M &= \frac{\sum_x \delta_x P (D_i = 1 | X_i = x) P (X_i = x)}{\sum_x P(D_i = 1 |X_i = x)P(X_i = x)} \\
&= \sum_x \delta_x \frac{ P (D_i = 1 | X_i = x) P (X_i = x)}{\sum_x P(D_i = 1 |X_i = x)P(X_i = x)}
\end{aligned}
$$

------------------------------------------------------------------------

### 회귀 추정대상

회귀에서는:

$$
Y_i = \sum_x d_{ix} \beta_x + \delta_R D_i + \varepsilon_i
$$

-   $d_{ix}$ = $X_i = x$임을 나타내는 지시변수
-   $\beta_x$ = $X = x$에서의 기초선 결과
-   $\delta_R$ = 회귀 추정대상

Then,

$$
\begin{aligned}
\delta_R &= \frac{\sum_x \delta_x [P(D_i = 1 | X_i = x) (1 - P(D_i = 1 | X_i = x))]P(X_i = x)}{\sum_x [P(D_i = 1| X_i = x)(1 - P(D_i = 1 | X_i = x))]P(X_i = x)} \\
&= \sum_x \delta_x \frac{[P(D_i = 1 | X_i = x) (1 - P(D_i = 1 | X_i = x))]P(X_i = x)}{\sum_x [P(D_i = 1| X_i = x)(1 - P(D_i = 1 | X_i = x))]P(X_i = x)}
\end{aligned}
$$

------------------------------------------------------------------------

### 해석: 가중치의 차이

매칭과 회귀의 차이는 공변량별 처치효과 $\delta_x$에 어떤 가중치를 부여하는지에 있다.

+-------------+--------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------+
| 유형        | 가중 함수                                                                   | 해석                                                                               | 타당한 이유                                                                                              |
+=============+======================================================================================+==============================================================================================+=====================================================================================================================+
| 매칭    | $P(D_i = 1 \mid X_i = x)$                                                            | 처치 단위가 많은 곳에 더 큰 가중치를 둔다(ATT 중심)                            | 처치집단의 효과가 관심 대상이므로 처치 단위가 관측되는 곳에 더 큰 가중치를 둔다            |
+-------------+--------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------+
| 회귀  | $\begin{aligned}P(D_i = 1 \mid X_i = x)\\(1 - P(D_i = 1 \mid X_i = x))\end{aligned}$ | 처치 배정의 분산이 높은 곳(즉, 처치/통제가 50/50에 가까운 곳)에 더 큰 가중치를 둔다 | 처치효과가 $X$ 전반에서 동질적이라고 가정하면 이러한 셀이 $\delta_x$의 최소 분산 추정치를 제공한다 |
+-------------+--------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------+

### 요약 표: 매칭과 회귀

+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| Feature                    | Matching                                                 | Regression (OLS)                                    |
+============================+==========================================================+=====================================================+
| **Functional Form**        | Less parametric; no assumption of linearity              | Parametric; usually assumes linearity               |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| **Primary Estimand**       | ATT (effect on the treated)                              | ATE or effects of continuous/interacted treatments  |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| **Balance**                | Enforces balance via matched samples                     | Does not guarantee balance                          |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| **Diagnostics**            | Covariate SMDs, QQ plots, empirical distributions        | Residual plots, R-squared, heteroskedasticity tests |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| **Unobserved Confounding** | Cannot be resolved; assumes ignorability                 | Same limitation                                     |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| **Standard Errors**        | Larger; require bootstrapping                            | Smaller; closed-form under assumptions              |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+
| **Best Used When**         | High control-to-treated ratio; misspecification concerns | Model is correctly specified; sufficient overlap    |
+----------------------------+----------------------------------------------------------+-----------------------------------------------------+

------------------------------------------------------------------------

### 정성적 비교

+---------------------------------------------------------------------+----------------------------------------------------------------------+
| Matching                                                            | Regression                                                           |
+=====================================================================+======================================================================+
| Not sensitive to the form of covariate-outcome relationship         | Can estimate continuous or interacted treatment effects              |
+---------------------------------------------------------------------+----------------------------------------------------------------------+
| Easier to assess balance and interpret diagnostics                  | Easier to estimate the effects of all covariates, not just treatment |
+---------------------------------------------------------------------+----------------------------------------------------------------------+
| Facilitates clear visual evaluation of overlap and balance          | Less intuitive diagnostics; model diagnostics used                   |
+---------------------------------------------------------------------+----------------------------------------------------------------------+
| Helps when treatment is rare (prunes clearly incomparable controls) | Performs better with balanced treatment assignment                   |
+---------------------------------------------------------------------+----------------------------------------------------------------------+
| Forces explicit enforcement of common support                       | May extrapolate outside the support of covariate distributions       |
+---------------------------------------------------------------------+----------------------------------------------------------------------+

------------------------------------------------------------------------

## 소프트웨어와 실제 구현

Many R packages provide functionality for implementing the various matching methods discussed above. Below is an overview of some popular options:

-   MatchIt:\
    Implements a wide range of matching methods (nearest neighbor, optimal, full, subclassification, exact, etc.). It focuses on "preprocessing" data before a final outcome analysis.

-   Matching:\
    Provides multivariate and propensity score matching, including options for *exact* and *nearest neighbor* matching. The package also offers functions to evaluate balance and to conduct sensitivity analyses.

-   cem ([Coarsened Exact Matching](#sec-cem)):\
    Uses a coarsening approach to create strata within which exact matching can be performed. This can reduce imbalance by discarding units that do not overlap in coarsened covariate space.

-   optmatch:\
    Enables *optimal matching* with variable matching ratios and full matching, allowing for flexible group constructions that minimize overall distance.

-   MatchingFrontier [@king2017balance]:\
    Finds the "frontier" of matching solutions by balancing sample size (or other constraints) against covariate balance. Allows analysts to see trade-offs in real time.

-   CBPS (Covariate Balancing Propensity Score):\
    Estimates propensity scores such that covariate balance is directly optimized. This can help avoid iterative re-specification of the propensity score model.

-   PanelMatch [@rauh2025panelmatch]:\
    Tailored to panel (longitudinal) data settings, providing matching methods that exploit repeated observations over time (e.g., for DID-type analyses in a time-series cross-sectional environment). Covered in depth, with a full replication and diagnostics, in Section \@ref(sec-panel-matching) at the end of this chapter.

-   PSAgraphics:\
    Specializes in visual diagnostics for propensity score analyses, offering graphical tools to inspect balance and common support.

-   rbounds:\
    Conducts Rosenbaum bounds sensitivity analysis on matched data. Researchers can examine how a hypothetical unmeasured confounder could undermine their estimated treatment effects.

-   twang:\
    Implements *generalized boosted models (GBM)* to estimate propensity scores. Often used for weighting approaches such as inverse probability weighting (IPW).

In practice, the choice of software and methods hinges on the study design, the nature of the data, and the researcher's theoretical expectations regarding treatment assignment.

------------------------------------------------------------------------

## 관측변수에 의한 선택 {#sec-selection-on-observables}

In observational studies, treatment assignment is typically not randomized. This poses a challenge when estimating causal effects, as treated and control groups may differ systematically. A central assumption that allows us to estimate causal effects from such data is selection on observables, also known as unconfoundedness or conditional independence.

Suppose we observe a binary treatment indicator $T_i \in {0, 1}$ and an outcome $Y_i$. Each unit $i$ has two potential outcomes:

-   $Y_i(1)$: outcome if treated
-   $Y_i(0)$: outcome if untreated

However, only one of these outcomes is observed for each unit. The average treatment effect on the treated (ATT) is:

$$
\text{ATT} = \mathbb{E}[Y(1) - Y(0) \mid T = 1]
$$

To identify this from data, we invoke the conditional independence assumption:

$$
(Y(0), Y(1)) \perp T \mid X
$$

This assumption means that after controlling for covariates $X$, treatment is as good as randomly assigned. A secondary assumption is overlap:

$$
0 < \mathbb{P}(T = 1 \mid X = x) < 1 \quad \text{for all } x
$$

This ensures that for every covariate profile, there is a positive probability of being both treated and untreated.

Matching attempts to approximate the conditions of a randomized experiment by creating a comparison group that is similar to the treated group in terms of observed covariates. Instead of relying solely on model-based adjustment (e.g., regression), matching balances the covariate distribution across treatment groups before estimation.

------------------------------------------------------------------------

### `MatchIt`을 사용한 매칭

We demonstrate the matching procedure using the `lalonde` dataset, a classic example in the causal inference literature, which investigates the effect of job training on subsequent earnings.

```{r}
library(MatchIt)
data("lalonde")
```

We focus on estimating the effect of the treatment (`treat`) on earnings in 1978 (`re78`), conditional on covariates.

------------------------------------------------------------------------

#### 1단계: 분석 계획

Before conducting matching, several strategic decisions must be made:

-   Estimand: do you want ATT (effect on treated), ATE (effect in population), or ATC (effect on controls)? Matching typically targets the ATT.

-   Covariate Selection: only include pre-treatment variables that are potential confounders, i.e., that affect both the treatment assignment and the outcome [@austin2011optimal; @vanderweele2019principles].

-   Distance Measure: choose how to quantify similarity between units (e.g., propensity score, Mahalanobis distance).

-   Matching Method: determine the method (e.g., nearest neighbor, full matching, genetic matching).

For our demonstration, we focus on ATT using propensity score matching.

------------------------------------------------------------------------

#### 2단계: 초기 불균형 평가

We first assess imbalance between treatment and control groups before matching.

```{r}
# Estimate propensity scores with logistic regression
m.out0 <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = MatchIt::lalonde,
  method = NULL,     # no matching, only estimates propensity scores
  distance = "glm"
)

# Summary of balance statistics before matching
summary(m.out0)
```

This summary provides:

-   Standardized mean differences

-   Variance ratios

-   Propensity score distributions

These diagnostics help us understand the extent of covariate imbalance.

------------------------------------------------------------------------

#### 3단계: 매칭 구현

1.  Nearest Neighbor Matching (1:1 without Replacement)

```{r}
m.out1 <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = MatchIt::lalonde,
  method = "nearest",
  distance = "glm"
)

```

Matching is based on estimated propensity scores. Each treated unit is matched to the closest control unit.

Assess Balance After Matching

The code below produces the post-matching balance diagnostics for the nearest-neighbour 1:1 propensity-score match: a jitter plot of the propensity-score distribution and QQ plots for selected covariates.

```{r mm-nearest-balance-diagnostics}
summary(m.out1, un = FALSE)  # only show post-matching stats

# Visual diagnostic: jitter plot of propensity scores
plot(m.out1, type = "jitter", interactive = FALSE)

# QQ plot for individual covariates
plot(m.out1, type = "qq", which.xs = c("age", "re74"))

```

**Interpretation**: Good matches will show overlapping distributions of covariates across groups, and standardized differences should be below 0.1 in absolute value.

2.  Full Matching

Allows many-to-one or one-to-many matches, minimizing overall distance.

```{r}
m.out2 <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = MatchIt::lalonde,
  method = "full",
  distance = "glm",
  link = "probit"
)

summary(m.out2, un = FALSE)

```

3.  Exact Matching

Only matches units with exactly the same covariate values (usually categorical):

```{r}
m.out3 <- matchit(
  treat ~ race + nodegree,
  data = MatchIt::lalonde,
  method = "exact"
)

```

4.  Optimal Matching

Minimizes the total distance between matched units across the sample.

```{r}
m.out4 <- matchit(
  treat ~ age + educ + re74 + re75,
  data = MatchIt::lalonde,
  method = "optimal",
  ratio = 2
)

```

5.  Genetic Matching

Searches over weights assigned to covariates to optimize balance.

```{r}
m.out5 <- matchit(
  treat ~ age + educ + re74 + re75,
  data = MatchIt::lalonde,
  method = "genetic"
)

```

------------------------------------------------------------------------

#### 4단계: 처치효과 추정

Once matching is complete, we use the matched data to estimate treatment effects.

```{r}
library(lmtest)
library(sandwich)

# Extract matched data
matched_data <- match.data(m.out1)

# Estimate ATT with robust standard errors
model <- lm(re78 ~ treat + age + educ + race + re74 + re75,
            data = matched_data,
            weights = weights)

coeftest(model, vcov. = vcovCL, cluster = ~subclass)

```

The coefficient on `treat` is our estimate of the ATT. The weights and subclass clustering account for the matched design.

### 보고 기준

To ensure transparency and reproducibility, always report the following:

1.  Matching method (e.g., nearest neighbor, genetic)

2.  Distance metric (e.g., propensity score via logistic regression)

3.  Covariates matched on, and justification for their inclusion

4.  Balance statistics (e.g., standardized mean differences before/after)

5.  Sample sizes: total, matched, unmatched, discarded

6.  Estimation model: whether treatment effects are estimated using regression adjustment, with or without weights

7.  Assumptions: especially unconfoundedness and overlap

------------------------------------------------------------------------

### `designmatch`를 통한 최적화 기반 매칭

For more advanced applications, the `designmatch` package provides matching methods based on combinatorial optimization.

```{r}
library(designmatch)
```

Notable methods:

-   `distmatch()`: Distance-based matching with custom constraints

-   `bmatch()`: Bipartite matching using linear programming

-   `cardmatch()`: Cardinality matching for maximum matched sample size with balance constraints

-   `profmatch()`: Profile matching for stratified treatment allocation

-   `nmatch()`: Non-bipartite matching (e.g., in interference settings)

------------------------------------------------------------------------

### MatchingFrontier

As mentioned in `MatchIt`, you have to make trade-off (also known as bias-variance trade-off) between balance and sample size. An automated procedure to optimize this trade-off is implemented in `MatchingFrontier` [@king2017balance], which solves this joint optimization problem.

Following `MatchingFrontier` [guide](https://projects.iq.harvard.edu/files/frontier/files/using_matchingfrontier.pdf)

```{r, message=FALSE, eval=FALSE}
# library(devtools)
# install_github('ChristopherLucas/MatchingFrontier')
library(MatchingFrontier)
data("lalonde", package = "MatchIt")
# choose var to match on
match.on <-
    colnames(lalonde)[!(colnames(lalonde) %in% c('re78', 'treat'))]
match.on

# Mahanlanobis frontier (default)
mahal.frontier <-
    makeFrontier(
        dataset = lalonde,
        treatment = "treat",
        match.on = match.on
    )
mahal.frontier

# L1 frontier
L1.frontier <-
    makeFrontier(
        dataset = lalonde,
        treatment = 'treat',
        match.on = match.on,
        QOI = 'SATT',
        metric = 'L1',
        ratio = 'fixed'
    )
L1.frontier

# estimate effects along the frontier

# Set base form
my.form <-
    as.formula(re78 ~ treat + age + black + education 
               + hispanic + married + nodegree + re74 + re75)

# Estimate effects for the mahalanobis frontier
mahal.estimates <-
    estimateEffects(
        mahal.frontier,
        're78 ~ treat',
        mod.dependence.formula = my.form,
        continuous.vars = c('age', 'education', 're74', 're75'),
        prop.estimated = .1,
        means.as.cutpoints = TRUE
    )

# Estimate effects for the L1 frontier
L1.estimates <-
    estimateEffects(
        L1.frontier,
        're78 ~ treat',
        mod.dependence.formula = my.form,
        continuous.vars = c('age', 'education', 're74', 're75'),
        prop.estimated = .1,
        means.as.cutpoints = TRUE
    )

# Plot covariates means 
# plotPrunedMeans()


# Plot estimates (deprecated)
# plotEstimates(
#     L1.estimates,
#     ylim = c(-10000, 3000),
#     cex.lab = 1.4,
#     cex.axis = 1.4,
#     panel.first = grid(NULL, NULL, lwd = 2,)
# )

# Plot estimates
plotMeans(L1.frontier)


# parallel plot
parallelPlot(
    L1.frontier,
    N = 400,
    variables = c('age', 're74', 're75', 'black'),
    treated.col = 'blue',
    control.col = 'gray'
)

# export matched dataset
# take 400 units
matched.data <- generateDataset(L1.frontier, N = 400) 
```

------------------------------------------------------------------------

### 성향점수 {#sec-propensity-scores}

Propensity score methods are widely used for estimating causal effects in observational studies, where random assignment to treatment and control groups is not feasible. Conceptually, the propensity score collapses a high-dimensional vector of covariates into a single scalar (the conditional probability of treatment), and matching, weighting, or stratifying on that scalar is sufficient to remove confounding from the *observed* covariates. The appeal is dimensional: instead of searching for control units that look similar across many variables at once, the analyst searches along a one-dimensional probability axis. The cost is that two units sharing the same propensity score can still differ markedly in covariate space, which is why propensity score methods trade balance on the joint distribution of $X$ for tractability. They are most defensible when the propensity model is well specified, the covariate set is rich enough to make [conditional ignorability](#sec-conditional-ignorability-assumption) credible, and [overlap](#sec-overlap-positivity-assumption) is strong.

The core idea is to mimic a randomized experiment by adjusting for confounding variables that predict treatment assignment. Formally, the propensity score is defined as the probability of assignment to treatment conditional on observed covariates [@rosenbaum1983central; @rosenbaum1985bias]:

$$
e_i(X_i) = P(T_i = 1 \mid X_i)
$$

where:

-   $T_i \in {0, 1}$ is the binary treatment indicator for unit $i$,
-   $X_i$ is a vector of observed pre-treatment covariates for unit $i$.

The key insight from @rosenbaum1983central is that conditioning on the propensity score is sufficient to remove bias due to confounding from observed covariates, under certain assumptions.

------------------------------------------------------------------------

#### 식별을 위한 가정

To identify causal effects using propensity scores, the following assumptions must hold:

-   Unconfoundedness / Conditional Independence Assumption (CIA):

$$
(Y_i(0), Y_i(1)) \perp T_i \mid X_i
$$

-   Positivity (Overlap):

$$
0 < P(T_i = 1 \mid X_i) < 1 \quad \text{for all } i
$$

These assumptions ensure that for each unit, we can observe comparable treated and untreated units in the sample. Violations of positivity, especially in high-dimensional covariate spaces, are a critical weakness of propensity score matching.

------------------------------------------------------------------------

#### PSM을 더 이상 권장하지 않는 이유

Despite its intuitive appeal, recent literature has strongly cautioned against using propensity score matching for causal inference [@king2019propensity]. The main criticisms are as follows:

-   Imbalance: PSM often fails to achieve covariate balance better than simpler techniques like covariate adjustment via regression or exact matching. Matching on the propensity score is a *scalar reduction* of a multivariate distribution, and this reduction can distort multivariate relationships.
-   Inefficiency: discarding unmatched units reduces statistical efficiency, especially when better estimators (e.g., inverse probability weighting or doubly robust estimators) can use all data.
-   Model dependence: small changes in the specification of the propensity score model can lead to large changes in matches and estimated treatment effects.
-   Bias: poor matches and irrelevant covariates can introduce additional bias rather than reduce it.

@abadie2016matching show that the asymptotic distribution of treatment effect estimators is sensitive to the estimation of the propensity score itself:

-   The estimated propensity score can improve efficiency over using the true propensity score when estimating the [ATE](#sec-average-treatment-effect). Formally, the adjustment to the asymptotic variance is non-positive.
-   However, for the [ATT](#sec-average-treatment-effect-on-the-treated), the sign of the adjustment is data-dependent. Estimation error in the propensity score can lead to misestimated confidence intervals: they may be too wide or too narrow.

This result suggests that even in large samples, failure to account for the estimation uncertainty in the propensity score can produce misleading inference.

A fundamental flaw in PSM is the asymmetry of match quality:

-   If $X_c = X_t$, then it must be that $e(X_c) = e(X_t)$.
-   However, the converse does not hold:\
    $e(X_c) = e(X_t) \nRightarrow X_c = X_t$

Therefore, two units with identical propensity scores may still differ substantially in covariate space. This undermines the matching goal of achieving similarity across all covariates.

------------------------------------------------------------------------

#### 성향점수 추정

Estimation of the propensity score is typically carried out using:

-   Parametric models:
    -   Logistic regression:\
        $$
        \hat{e}_i = \frac{1}{1 + \exp(-X_i^\top \hat{\beta})}
        $$
-   Nonparametric / machine learning methods:
    -   Generalized Boosted Models (GBM)
    -   Boosted Classification and Regression Trees (CART)
    -   Random forests or Bayesian Additive Regression Trees (BART)

These machine learning approaches often yield better balance due to flexible functional forms, but they also complicate interpretation and inference.

------------------------------------------------------------------------

#### 매칭 알고리즘

1.  Reduce the high-dimensional vector $X_i$ to the scalar $\hat{e}_i$.
2.  Calculate distances between treated and control units based on $\hat{e}_i$: $$
    d(i, j) = |\hat{e}_i - \hat{e}_j|
    $$
3.  Match each treated unit $i$ to the control unit $j$ with the closest propensity score.
4.  Prune:
    -   Control units not used in any match are discarded.
    -   Matches exceeding a caliper (maximum distance threshold) are also discarded.
5.  No replacement is typically assumed; each control unit is matched at most once.

Let $c > 0$ be a caliper. Then a treated unit $i$ is matched to control unit $j$ only if:

$$
|\hat{e}_i - \hat{e}_j| < c
$$

Caliper matching helps reduce poor matches, but may result in *random pruning*, further reducing balance and efficiency.

------------------------------------------------------------------------

#### 실무 권고사항

-   Do not include irrelevant covariates: including variables that are unrelated to the outcome can increase the variability of the estimated propensity score and reduce matching quality.
-   Avoid instrumental variables [@bhattacharya2007instrumental]: including IVs in the propensity score model can inflate bias by introducing variation in treatment assignment unrelated to potential outcomes.
-   Focus on covariates that are confounders, i.e., those that affect both treatment and outcome.

What remains after pruning is more consequential than the initial covariate set. Matching can discard a large portion of the sample, which distorts representativeness and increases variance.

------------------------------------------------------------------------

#### 진단과 평가

After matching, the primary diagnostic tool is covariate balance. Key diagnostics include:

-   **Standardized mean differences (SMD)** between treatment groups before and after matching
-   **Variance ratios**
-   **Visual inspection** via Love plots or density plots

Statistical model fit criteria (e.g., AIC, BIC, c-statistics) are **not valid** for evaluating propensity score models, since the goal is *not prediction*, but achieving balance.

There is **no need to worry about collinearity** in the covariates when estimating propensity scores --- unlike in outcome regression models, where multicollinearity can inflate standard errors.

------------------------------------------------------------------------

#### 비즈니스와 금융 분야의 응용

Propensity score methods have been used in empirical finance and marketing, though increasingly replaced by more robust approaches. One illustrative application is found in @hirtle2020impact, which investigates the causal effect of regulatory bank supervision on firm-level outcomes:

-   Treatment: Degree of supervisory attention.
-   Outcomes: Loan risk, profitability, volatility, and firm growth.
-   Method: Propensity score matching to construct treated and control groups of banks with comparable observed characteristics.

Their matched sample analysis reveals that **intensified supervision** leads to:

-   Lower risk (more conservative loan portfolios)
-   Reduced volatility
-   No significant loss in profitability or growth

------------------------------------------------------------------------

#### 결론

While the theoretical motivation behind propensity scores remains sound, their application via naive matching methods is no longer considered best practice. The statistical community increasingly favors alternatives such as:

-   **Covariate adjustment via regression**
-   **Inverse probability weighting (IPW)**
-   **Doubly robust estimators**
-   **Targeted Maximum Likelihood Estimation (TMLE)**

These approaches better utilize the full dataset, yield more efficient estimators, and offer more transparent diagnostics. Matching on estimated propensity scores may still be useful for illustration or sensitivity analysis, but should not be the primary method for causal inference in applied research.

------------------------------------------------------------------------

#### 미래 정보를 이용한 성향점수 매칭

In observational data, estimating **causal effects** is complicated by **self-selection bias**: individuals who receive a treatment may differ systematically from those who do not. Standard **PSM** methods attempt to control for such biases by matching treated and untreated units on observable characteristics.

However, when **unobservable** or **latent traits** influence both treatment assignment and outcomes, traditional PSM may produce biased estimates. @bapna2018monetizing introduce **Look-Ahead Propensity Score Matching (LA-PSM)**, a novel technique that leverages **future behavior** to correct for **time-invariant unobserved confounding**, particularly useful in rare-event economic decisions.

In the study of @bapna2018monetizing:

-   Treatment: user pays for a premium subscription.
-   Outcome: social engagement (songs listened, playlists created, friends made).
-   Problem: users choosing to subscribe are systematically different (e.g., more engaged).

##### LA-PSM Implementation

-   Treated group: users subscribing between Sept 2011 - June 2012.

-   Control group: users who will subscribe later (June 2012 - Jan 2015).

-   Matching on: observed demographics, past behavior.

-   Analysis: [Difference-in-Differences](#sec-difference-in-differences) estimation.

**Results:** Premium adoption increases:

-   Songs listened: +287.2%
-   Playlists created: +1.92%
-   Forum posts: +2.01%
-   Friends added: +15.77%

------------------------------------------------------------------------

In standard PSM:

-   Match individuals who received treatment with individuals who did not, based on observed covariates.
-   Assume selection into treatment is **strongly ignorable**, conditional on covariates.

Problem:

-   Unobserved factors (e.g., ambition, risk tolerance) may bias both treatment and outcome.

-   Standard PSM cannot correct for this.

##### Solution: Look-Ahead PSM

-   Match treated units with **future-treated** units: those who haven't yet received treatment but will in the future.

-   Future-treated units share unobservable traits with current-treated units.

Thus, LA-PSM controls for **time-invariant unobserved confounders**.

------------------------------------------------------------------------

Let:

-   $i \in \{1, \ldots, N\}$ index individuals.

-   $D_i(t)$ = 1 if individual $i$ is treated at time $t$, 0 otherwise.

-   $X_i$ = observed covariates.

-   $U_i$ = unobserved time-invariant confounders.

-   $Y_i(t)$ = outcome at time $t$.

In standard PSM:

-   Match on $\mathbb{P}(D_i(t) = 1 \mid X_i)$.

In LA-PSM:

-   Match **current treated** individuals with **future treated** individuals based on: $$ \mathbb{P}(D_i(t) = 1 \mid X_i) $$

-   and require: $$ \exists  s > t \quad \text{such that} \quad D_i(s) = 1 $$

Formally, define:

-   Treatment group: $$ T = \{ i \mid D_i(t) = 1 \} $$

-   Control group: $$ C = \{ j \mid D_j(t) = 0 \quad \text{and} \quad \exists  s>t : D_j(s) = 1 \} $$

Thus, both treatment and control groups are "eventual adopters," just at different times.

------------------------------------------------------------------------

**Proposition:**\
If $U_i$ is **time-invariant** and **affects both treatment assignment and outcomes**, then LA-PSM produces unbiased causal estimates under weaker assumptions than standard PSM.

1.  In regular PSM, matching on $X_i$ cannot adjust for $U_i$.
2.  In LA-PSM, by restricting controls to **future adopters**, we select units that share latent traits $U_i$ driving adoption.
3.  Thus, $U_i$ is balanced across treatment and control, eliminating bias from $U_i$.
4.  Remaining bias is due only to differences in $X_i$, which are adjusted for via matching.

------------------------------------------------------------------------

##### Practical Implementation Steps

1.  In **Static Look-Ahead PSM**, we:

-   Fix a single matching window at a given time $t$.
-   Define:
    -   Treatment Group: Individuals who receive treatment at time $t$.
    -   Control Group: Individuals who are not yet treated at $t$ but will receive treatment at a later time $t'>t$.
-   Estimate propensity scores based only on observed covariates $X$.
-   Match treated and (future-treated) control units based on their propensity scores.

This ensures treated and control individuals have similar observed $X$ and similar unobserved $U$, assuming $U$ is time-invariant.

How to implement Static LA-PSM:

1.  Identify individuals **treated now** (treatment group) and **future adopters** (control group).
2.  Estimate propensity scores $\mathbb{P}(D=1 \mid X)$.
3.  Match treated and control units using nearest-neighbor matching.
4.  Estimate treatment effects via a regression on the matched sample.

Important: 

-   Drop individuals who are never treated from the analysis.

-   Only compare current-treated and future-treated!

------------------------------------------------------------------------

2.  In **Dynamic Look-Ahead PSM**, we:

-   Allow time to move forward ($t=1,2,3,\dots$).
-   At each time $t$:
    -   Define treated individuals (those who receive treatment at $t$).
    -   Define control individuals (those untreated at $t$ but who will adopt later).
-   Repeat matching separately for each $t$.
-   Aggregate results across different $t$ periods to estimate the overall effect.

Dynamic LA-PSM is more flexible:

-   It updates control groups over time.

-   Better handles situations where treatment adoption spreads gradually over time.

How to implement Dynamic LA-PSM:

1.  For each time period $t$:
    -   Define treated-now = individuals treated at $t$.
    -   Define control = individuals untreated at $t$ but who adopt later.
2.  Estimate propensity scores $\mathbb{P}(D=1 \mid X)$ within that time window.
3.  Match treated and controls at each $t$ separately.
4.  Pool matched samples from all $t$ together.
5.  Estimate treatment effects using the pooled sample.

Important: 

-   Keep track of time --- matching is done separately each period.

-   Aggregate the treatment effect estimates carefully (either by averaging or pooling matched data).

------------------------------------------------------------------------

+--------------------+------------------------------------+--------------------------------------------+
| Feature            | Static Look-Ahead PSM              | Dynamic Look-Ahead PSM                     |
+:===================+:===================================+:===========================================+
| Matching Window    | Single, fixed window (e.g., $t=5$) | Rolling window at each time $t$            |
+--------------------+------------------------------------+--------------------------------------------+
| Treated Group      | Treated at $t$                     | Treated at $t$                             |
+--------------------+------------------------------------+--------------------------------------------+
| Control Group      | Future treated (after $t$)         | Future treated relative to each $t$        |
+--------------------+------------------------------------+--------------------------------------------+
| Updates over time? | No                                 | Yes                                        |
+--------------------+------------------------------------+--------------------------------------------+
| Suitable for       | Simple adoption settings           | Gradual adoption / time-sensitive settings |
+--------------------+------------------------------------+--------------------------------------------+

> **Challenge:**\
> Simulate your own dataset with hidden confounding and implement both Static and Dynamic Look-Ahead PSM.\
> Compare bias against Standard PSM and Randomized Assignment.

------------------------------------------------------------------------

### Mahalanobis 거리 매칭 {#sec-mahalanobis}

Where [propensity score matching](#sec-propensity-scores) compresses covariates into a single scalar, Mahalanobis Distance Matching takes the opposite tack: it works in the full multivariate covariate space and asks which control units are jointly close to each treated unit on every covariate at once. The trade-off is straightforward. Mahalanobis matching tends to deliver better joint covariate balance when it succeeds, because two units close in Mahalanobis distance are close on each variable individually; but as the number of covariates grows, "close on every variable at once" becomes scarce, and matches deteriorate. In the bias-variance language of the previous section, Mahalanobis matching pushes hard on bias reduction (assuming the matches succeed) and pays for it in sample retention as dimensionality grows.

Mahalanobis Distance Matching is a method for matching units in observational studies based on the **multivariate similarity** of covariates. Unlike propensity score matching, which reduces the covariate space to a single scalar, Mahalanobis distance operates in the full multivariate space. As a result, Mahalanobis matching can be interpreted as approximating a **fully blocked design**, where each treated unit is paired with a control unit that has nearly identical covariate values.

-   In the ideal case: $X_t = X_c$, implying a perfect match.
-   In practice, Mahalanobis distance allows for "near-exact" matches in high-dimensional space.

Because it preserves the multivariate structure of the covariates, Mahalanobis matching more faithfully emulates randomization within **covariate strata**, making it more robust to specification error than propensity score matching.

This method is particularly appealing when the number of covariates is relatively small and when these covariates are continuous and well-measured.

------------------------------------------------------------------------

Given a set of $p$ covariates $X_i \in \mathbb{R}^p$ for unit $i$, the **Mahalanobis distance** between a treated unit $t$ and a control unit $c$ is defined as:

$$
D_M(X_t, X_c) = \sqrt{(X_t - X_c)^\top S^{-1}(X_t - X_c)}
$$

where:

-   $X_t$ and $X_c$ are the $p$-dimensional vectors of covariates for treated and control units, respectively,
-   $S$ is the sample covariance matrix of the covariates across all units (treated and control),
-   $S^{-1}$ serves to standardize and decorrelate the covariate space, accounting for both scale and correlation among covariates.

------------------------------------------------------------------------

#### 유클리드 거리를 사용하지 않는 이유

The Euclidean distance:

$$
D_E(X_t, X_c) = \sqrt{(X_t - X_c)^\top (X_t - X_c)}
$$

does **not adjust for different variances** among covariates or for **correlation** between them. Mahalanobis distance corrects for this by incorporating the inverse covariance matrix $S^{-1}$, effectively transforming the data to a space where the covariates are **standardized and orthogonal**.

This makes the Mahalanobis distance **scale-invariant**, i.e., invariant under affine transformations of the data, which is essential when matching on variables of different units (e.g., income in dollars and age in years).

------------------------------------------------------------------------

#### Mahalanobis 매칭 알고리즘

Let $\mathcal{T}$ be the set of treated units and $\mathcal{C}$ the set of control units. The procedure for Mahalanobis matching can be described as follows:

1.  **Compute the covariance matrix** $S$ of the covariates $X$ across all units.
2.  **For each treated unit** $i \in \mathcal{T}$, compute $D_M(X_i, X_j)$ for all $j \in \mathcal{C}$.
3.  **Match** treated unit $i$ to the control unit $j$ with the smallest Mahalanobis distance.
4.  Prune: 
    -   Control units not used in any match are discarded.
    -   Matches where $D_M > \delta$ (a caliper threshold) are also discarded.

A caliper $\delta$ is used to enforce a maximum allowable distance. That is, a match is made between $i$ and $j$ only if:

$$
D_M(X_i, X_j) < \delta
$$

This prevents poor-quality matches and helps ensure that treated and control units are meaningfully similar. If no match falls within the caliper for a given treated unit, that unit is left unmatched, and potentially discarded from the analysis.

------------------------------------------------------------------------

#### 특성

-   Scale-invariant: Standardizes variables using $S^{-1}$, ensuring that variables with large scales do not dominate the distance metric.
-   Correlation-adjusted: Accounts for linear relationships among covariates, which is critical in multivariate contexts.
-   Non-parametric: No model for treatment assignment is estimated; purely based on observed covariates.

------------------------------------------------------------------------

#### 한계

-   Sensitive to multicollinearity: If the covariates are highly collinear, the covariance matrix $S$ may be nearly singular, making $S^{-1}$ unstable or non-invertible. Regularization techniques may be required.
-   Not suitable for high-dimensional covariate spaces: As $p$ increases, exact or even near-exact matches become harder to find. Dimensionality reduction techniques (e.g., PCA) may help.
-   Inefficient with categorical variables: Since Mahalanobis distance is based on continuous covariates and assumes multivariate normality (implicitly), it's less effective when most covariates are categorical or binary.

------------------------------------------------------------------------

#### 하이브리드 접근법

Mahalanobis matching can be **combined with propensity scores** for improved performance:

-   Within propensity score calipers: Match using Mahalanobis distance only within groups of units that fall within a specified caliper of each other in propensity score space.
-   Stratified matching: Divide the sample into strata based on propensity scores and apply Mahalanobis matching within each stratum.

These hybrid methods aim to preserve the robustness of Mahalanobis matching while mitigating its weaknesses in high dimensions.

------------------------------------------------------------------------

#### 실무 고려사항

-   Estimate $S$ from the pooled sample (treated + control) to ensure consistency.
-   Standardize all covariates before applying Mahalanobis distance to ensure comparability and numerical stability.
-   Use diagnostic plots (e.g., QQ plots or multivariate distance histograms) to assess match quality.

------------------------------------------------------------------------

### 구간화 정확 매칭(CEM) {#sec-cem}

Coarsened Exact Matching addresses a different question than [propensity score](#sec-propensity-scores) or [Mahalanobis](#sec-mahalanobis) matching. Rather than letting an algorithm decide what counts as "close," CEM asks the analyst to declare it: how finely should age be binned? Should education be split into four categories or eight? Once the analyst makes those substantive choices, exact matching within the resulting strata is straightforward, and the maximum imbalance is bounded by construction. This makes CEM transparent and tunable, with the trade-off being explicit. Coarser bins keep more of the sample but allow more residual imbalance within each stratum; finer bins guarantee tight balance but discard strata that lack both treated and control units. CEM is often the right choice when the covariate set is moderate, when the analyst has substantive intuition about meaningful cut points, and when a non-parametric procedure that does not lean on a treatment-assignment model is preferred.

Coarsened Exact Matching (CEM) is a **monotonic imbalance bounding** matching method designed to improve covariate balance between treated and control units in observational studies. CEM operates by **coarsening** continuous or high-cardinality covariates into discrete bins and then applying **exact matching** on this coarsened space. The result is a non-parametric method that prioritizes covariate balance and robustness over modeling assumptions.

Introduced and formalized in @iacus2012causal, CEM is well-suited for causal inference when covariates are noisy or when traditional modeling-based approaches (e.g., propensity scores) are unstable or hard to interpret.

------------------------------------------------------------------------

The central idea is to **discretize covariates** into meaningful bins (either automatically or manually), then sort observations into **strata** or subclasses based on the **unique combinations** of these binned covariate values. Exact matching is applied within each stratum.

Let $X_i \in \mathbb{R}^p$ be the $p$-dimensional covariate vector for unit $i$. Define $C(X_i)$ to be a coarsened version of $X_i$, where:

-   Continuous variables are binned into intervals (e.g., age 20--29, 30--39, etc.)
-   Categorical variables may be grouped (e.g., Likert scales aggregated into fewer categories)

Then, matching proceeds as follows:

1.  Temporarily coarsen the covariate space $X \to C(X)$.
2.  Sort observations into strata defined by unique values of $C(X)$.
3.  Prune any stratum that contains only treated or only control units.
4.  Retain all original (uncoarsened) units in matched strata for analysis.

This process produces a matched sample in which each stratum includes at least one treated and one control unit, improving internal validity by design.

As with other matching methods, CEM requires the **ignorability assumption** (also known as unconfoundedness or selection on observables):

$$ (Y_i(0), Y_i(1)) \perp T_i \mid X_i $$

Under this assumption and sufficient overlap in the coarsened strata, CEM yields unbiased estimates of causal effects within the matched sample.

------------------------------------------------------------------------

#### Mathematical Properties

1.  **Monotonic Imbalance Bounding**

CEM is a Monotonic Imbalance Bounding method, meaning that the user can pre-specify the maximum level of imbalance allowed on each covariate. The imbalance measure is guaranteed to be weakly decreasing as coarsening becomes finer. This provides:

-   Transparency: The imbalance tradeoff is determined *ex ante* by the user.
-   Control: Users can make a direct decision about how much imbalance is tolerable.

Formally, let $\mathcal{L}(C(X))$ denote a measure of imbalance under coarsened covariates. Then, for any refinements of $C(X)$:

$$
\text{if } C_1(X) \preceq C_2(X), \text{ then } \mathcal{L}(C_1(X)) \leq \mathcal{L}(C_2(X))
$$

That is, finer coarsenings (more bins) cannot increase imbalance.

2.  **Congruence Principle**

CEM respects the congruence principle, which asserts that analysis should not be more precise than the data allow. By coarsening, CEM protects against overfitting and artificial precision, especially when measurement error is present.

3.  **Robustness**

-   Robust to measurement error: Discretization makes matching less sensitive to noise in covariates.
-   Works with missing data: The `cem` R package supports partial handling of missingness.
-   Multiple imputation: CEM can be applied within imputation routines to preserve matching structure across imputed datasets.
-   Supports multi-valued treatments: Matching can be extended to treatments beyond binary $T \in {0, 1}$.

------------------------------------------------------------------------

1.  Load and Prepare Data

```{r, message=FALSE}
library(cem)
data(LeLonde)

# Remove missing values
Le <- na.omit(LeLonde)

# Treatment and control indices
tr <- which(Le$treated == 1)
ct <- which(Le$treated == 0)
ntr <- length(tr)
nct <- length(ct)

# Unadjusted bias (naïve difference in means)
mean(Le$re78[tr]) - mean(Le$re78[ct])
```

2.  Define Pre-Treatment Covariates

```{r}
vars <- c(
    "age", "education", "black", "married", "nodegree",
    "re74", "re75", "hispanic", "u74", "u75", "q1"
)

# Pre-treatment imbalance
imbalance(group = Le$treated, data = Le[vars]) # L1 imbalance = 0.902

```

3.  Automatically Coarsen and Match

```{r}
mat <- cem(
    treatment = "treated",
    data = Le,
    drop = "re78",     # outcome variable
    keep.all = TRUE    # retain unmatched units in output
)
mat

```

-   `mat$w` contains **weights** for matched units.

-   Summary of matched strata and units retained.

4.  Manual Coarsening (User-Controlled)

Users may coarsen variables explicitly based on theoretical knowledge or data distribution.

Example: Grouped Categorical and Binned Continuous Covariates

```{r}
# Inspect levels for grouping
levels(Le$q1)

# Group Likert responses
q1.grp <- list(
    c("strongly agree", "agree"),
    c("neutral", "no opinion"),
    c("strongly disagree", "disagree")
)

# Custom cutpoints for education
table(Le$education)
educut <- c(0, 6.5, 8.5, 12.5, 17)

# Run CEM with manual coarsening
mat1 <- cem(
    treatment = "treated",
    data = Le,
    drop = "re78",
    cutpoints = list(education = educut),
    grouping = list(q1 = q1.grp)
)
mat1

```

This allows for **domain-specific discretion** and enhances interpretability.

#### 점진적 구간화

CEM supports **progressive coarsening**, where matching is attempted with fine coarsening first. If insufficient matches are found, coarsening is gradually relaxed until a suitable number of matched units is retained.

This strategy balances:

-   **Precision** (finer bins)

-   **Sample size retention** (coarser bins)

#### 요약

Strengths

-   Transparent and tunable matching procedure

-   Non-parametric: Does not require estimation of a treatment assignment model

-   Robust to measurement error and model misspecification

-   Supports multi-valued treatments and partial missingness

-   Theoretically grounded via monotonic imbalance bounding

Limitations

-   Loss of information due to coarsening

-   Arbitrary binning may influence results if not theoretically motivated

-   Not designed for high-dimensional settings without careful variable selection

-   Cannot account for unobserved confounding

Coarsened Exact Matching offers a powerful, transparent, and theoretically principled method for preprocessing observational data before estimating causal effects. It overcomes several limitations of traditional matching approaches, particularly propensity score matching, by giving researchers direct control over the quality of matches and by anchoring inference in empirical balance rather than model-dependent estimates.

------------------------------------------------------------------------

### 유전 알고리즘 매칭 {#sec-genetic-matching}

Genetic Matching extends the algorithms above by recognizing that the choice of distance metric is itself a researcher degree of freedom. Standard [Mahalanobis distance](#sec-mahalanobis) implicitly assigns weights via the inverse covariance matrix, which is a sensible default but is not optimized for balance on the specific covariates the analyst cares about. Genetic Matching instead searches over the space of weight matrices to find the one that best balances a user-specified set of covariates and transformations, using a stochastic optimizer modeled on natural selection. The method shines when conventional distance metrics leave residual imbalance on substantively important covariates, and when the analyst is willing to pay the computational cost of a population-based search. The trade-offs are familiar: relative to nearest-neighbor matching on a propensity score, Genetic Matching usually delivers better balance and lower bias, at the price of longer run times and an estimand that depends on which covariates the analyst chose to balance.

Genetic Matching (GM) is a generalization of [propensity score](#sec-propensity-scores) and [Mahalanobis distance matching](#sec-mahalanobis), leveraging a search algorithm inspired by evolutionary biology to optimize covariate balance. Unlike classical matching techniques, which rely on pre-specified distance metrics, Genetic Matching *learns* an optimal distance metric through an iterative process aimed at minimizing imbalance between treated and control groups.

Genetic Matching was introduced by @diamond2013genetic and is designed to improve balance on observed covariates by selecting optimal weights for each covariate. The core idea is to define a generalized Mahalanobis distance metric, where the weights used in the distance calculation are chosen by a **genetic search algorithm**. This process adaptively explores the space of possible weighting matrices and evolves toward a set of weights that result in optimal covariate balance across treatment groups.

The method combines two components:

-   [**Propensity Score Matching**](#sec-propensity-scores): Balances on the estimated probability of treatment assignment given covariates.
-   [**Mahalanobis Distance Matching**](#sec-mahalanobis): Accounts for correlations among covariates and ensures geometric proximity in multivariate space.

The Genetic Matching approach treats the selection of covariate weights as an optimization problem, solved using techniques inspired by natural selection: mutation, crossover, and survival of the fittest.

Compared to traditional matching methods such as:

-   Nearest Neighbor Matching: May result in poor balance in high-dimensional or imbalanced datasets.
-   Full Matching: More efficient but not always feasible with severe imbalance.

Genetic Matching is particularly powerful in situations where traditional distance metrics are insufficient to achieve balance. It adaptively searches the weight space to ensure better covariate overlap, which is crucial for unbiased causal inference.

In empirical settings, this often results in superior balance on observed covariates and more credible estimates of treatment effects.

------------------------------------------------------------------------

Let:

-   $T_i \in {0, 1}$ be the binary treatment indicator for unit $i$.

-   $\mathbf{X}_i \in \mathbb{R}^p$ be the covariate vector for unit $i$, where $p$ is the number of observed covariates.

-   $\mathbf{W}$ be a $p \times p$ positive definite weight matrix optimized by the algorithm.

The **generalized Mahalanobis distance** between unit $i$ and $j$ is defined as:

$$
D_{ij} = \sqrt{(\mathbf{X}_i - \mathbf{X}_j)^\top \mathbf{W} (\mathbf{X}_i - \mathbf{X}_j)}
$$

Here, $\mathbf{W}$ is not necessarily the inverse of the covariance matrix (as in standard Mahalanobis distance), but a data-driven weighting matrix found by Genetic Matching to optimize covariate balance.

The genetic algorithm optimizes the weights in $\mathbf{W}$ to minimize a global imbalance metric $\mathcal{B}$ across all covariates. This can involve different balance criteria:

-   Paired $t$-tests for continuous or dichotomous variables
-   Kolmogorov--Smirnov (K--S) test statistics for distributional similarity
-   Standardized mean differences

Let $b_k$ denote the balance metric for the $k$th covariate. Then the total imbalance could be:

$$
\mathcal{B} = \sum_{k=1}^{p} w_k b_k^2
$$

The optimization objective becomes finding $\mathbf{W}$ that minimizes $\mathcal{B}$.

------------------------------------------------------------------------

Genetic Matching is implemented in the `Matching` package in R via the `GenMatch()` function. This function:

-   Accepts a treatment indicator (`Tr`) and a matrix of covariates (`X`).
-   Optionally takes a **Balance Matrix** (`BalanceMatrix`) that specifies which variables should be balanced.
-   Uses a genetic search to find the weight matrix that best achieves balance on the specified variables.

Balance is assessed via:

-   Paired $t$-tests for dichotomous or continuous variables
-   Kolmogorov-Smirnov (K--S) tests for distributional differences

Matching can be performed:

-   **With or without replacement** (with replacement often improves match quality at the cost of increased variance).
-   For different estimands: [Average Treatment Effect], [Average Treatment Effect on the Treated], or [Average Treatment Effect on the Control].

------------------------------------------------------------------------

```{r, eval=FALSE}
library(Matching)
data(lalonde, package = "MatchIt")
attach(lalonde)

# Define the covariates to match on
X <- cbind(age, educ, black, hisp, married, nodegr, u74, u75, re75, re74)

# Define the covariates to balance on
BalanceMat <- cbind(
  age, educ, black, hisp, married, nodegr,
  u74, u75, re75, re74,
  I(re74 * re75) # Include interaction term
)

# Genetic Matching to optimize covariate balance
# Note: pop.size = 16 is too small for real applications
genout <- GenMatch(
  Tr = treat,
  X = X,
  BalanceMatrix = BalanceMat,
  estimand = "ATE",
  M = 1,
  pop.size = 16,
  max.generations = 10,
  wait.generations = 1
)

# Define the outcome variable
Y <- re78 / 1000

# Perform matching with the optimized weights
mout <- Match(
  Y = Y,
  Tr = treat,
  X = X,
  estimand = "ATE",
  Weight.matrix = genout
)

# Summarize the treatment effect estimates
summary(mout)

# Assess post-matching balance
mb <- MatchBalance(
  treat ~ age + educ + black + hisp + married + nodegr +
    u74 + u75 + re75 + re74 + I(re74 * re75),
  match.out = mout,
  nboots = 500
)
```

#### 확장과 고려사항

-   The method can be extended to multinomial treatments (via generalized entropy balancing).

-   For longitudinal or panel data, entropy balancing can be adapted to match across time points.

-   One should avoid balancing on post-treatment variables, which would bias estimates.

Entropy balancing fits naturally into modern workflows for causal inference, and is especially valuable when researchers want to prioritize design over modeling.

------------------------------------------------------------------------

### 엔트로피 균형

Entropy balancing departs from the matching paradigm in an instructive way. Rather than picking which control units to keep and which to drop, it keeps every control unit and instead chooses weights that, by construction, force the reweighted control group to match the treated group on a set of pre-specified moments (means, variances, and so on). Because no unit is discarded, the method is attractive when treated units are scarce or when sample size cannot be sacrificed. Because balance is achieved exactly on the chosen moments, no iterative diagnostic loop is needed. The trade-offs run in two directions. Entropy balancing avoids the model dependence of [propensity score](#sec-propensity-scores) estimation but transfers that dependence to the analyst's choice of which moments to balance: if a relevant interaction or non-linear function of covariates is omitted from the constraints, residual imbalance there can still bias the effect estimate. Extreme weights, while less common than under inverse probability weighting, can still inflate variance.

Entropy balancing is a data preprocessing technique for causal inference in observational studies. Introduced by @hainmueller2012entropy, it is particularly useful in settings with binary treatments where covariate imbalance between treated and control groups threatens the validity of causal estimates.

The method applies a **maximum entropy reweighting** scheme to assign **unit weights** to control group observations such that the **reweighted covariate distribution** in the control group matches the **sample moments** (e.g., means, variances) of the treated group.

Entropy balancing directly targets covariate balance and avoids the trial-and-error of iterative propensity score or matching methods. It also reduces reliance on outcome models by ensuring pre-treatment covariates are aligned between groups.

-   Goal: Create a set of weights for the control group such that its covariate distribution matches that of the treated group.
-   Approach: Solve a constrained optimization problem that minimizes information loss (measured via Shannon entropy) subject to **balance constraints** on covariates.

Entropy balancing:

-   Achieves exact balance on the specified covariate moments (e.g., means, variances).

-   Produces unique, data-driven weights without the need for iterative matching.

-   Is compatible with any outcome model in the second stage (e.g., weighted regression, IPW, DID, etc.).

#### 엔트로피 균형의 장점

-   Exact balance: Guarantees moment balance without iterative diagnostics.
-   Flexibility: Can balance on higher moments, interactions, or non-linear transformations.
-   Model independence: Reduces reliance on correct specification of the outcome model.
-   Stability: The optimization procedure is smooth and produces stable, interpretable weights.

Entropy balancing is especially useful in high-dimensional settings or when working with small treated groups, where matching can perform poorly or fail entirely.

------------------------------------------------------------------------

Suppose we have $n_1$ treated units and $n_0$ control units, with covariates $\mathbf{X}_i \in \mathbb{R}^p$ for each unit $i$. Let $T_i \in {0,1}$ be the treatment indicator.

Let the treatment group's sample moment vector be:

$$
\bar{\mathbf{X}}_T = \frac{1}{n_1} \sum_{i: T_i=1} \mathbf{X}_i
$$

We seek a set of weights $\{w_i\}_{i: T_i=0}$ for control units such that:

1.  **Covariate balancing constraints** (e.g., mean balance):

$$
\sum_{i: T_i=0} w_i \mathbf{X}_i = \bar{\mathbf{X}}_T
$$

2.  **Minimum entropy divergence** from uniform weights:

We minimize the Kullback-Leibler divergence between the new weights $w_i$ and uniform base weights $w_i^{(0)} = \frac{1}{n_0}$:

$$
\min_{\{w_i\}} \sum_{i: T_i=0} w_i \log \left( \frac{w_i}{w_i^{(0)}} \right)
$$

subject to:

-   $\sum w_i = 1$ (weights must sum to 1)
-   $\sum w_i \mathbf{X}_i = \bar{\mathbf{X}}_T$
-   (Optionally) higher-order moments, e.g., variances, skewness, etc.

This is a **convex optimization problem**, ensuring a unique global solution.

------------------------------------------------------------------------

```{r, eval = FALSE}
# Load package
library(ebal)

# Simulate data
set.seed(123)
n_treat <- 50
n_control <- 100

# Covariates
X_treat <- matrix(rnorm(n_treat * 3, mean = 1), ncol = 3)
X_control <- matrix(rnorm(n_control * 3, mean = 0), ncol = 3)
X <- rbind(X_control, X_treat)  # Order: control first, then treated

# Treatment vector: 0 = control, 1 = treated
treatment <- c(rep(0, n_control), rep(1, n_treat))

# Apply entropy balancing
eb_out <- ebalance(Treatment = treatment, X = X)

# Simulate outcome variable
Y_control <- rnorm(n_control, mean = 1)
Y_treat <- rnorm(n_treat, mean = 3)
Y <- c(Y_control, Y_treat)

# Construct weights: treated get weight 1, control get weights from ebalance
weights <- c(eb_out$w, rep(1, n_treat))

# Estimate ATE using weighted linear regression
df <- data.frame(Y = Y, treat = treatment, weights = weights)
fit <- lm(Y ~ treat, data = df, weights = weights)
summary(fit)
```

------------------------------------------------------------------------

### 고차원 자료의 매칭

As the dimensionality of covariates increases, traditional matching methods (such as nearest neighbor matching using propensity scores or Mahalanobis distance) become increasingly unreliable. This issue, often referred to as the **curse of dimensionality**, undermines the ability to find good matches because distances between points become less informative in high-dimensional space.

In high-dimensional settings, where the number of covariates is large (potentially exceeding the number of observations), careful preprocessing is necessary to reduce dimensionality before matching can be effectively applied. The following approaches are commonly used to mitigate these challenges by reducing the feature space while preserving meaningful structure relevant to treatment assignment and outcomes.

#### 차원 축소 기법

A variety of dimensionality reduction techniques can be employed prior to matching. The methods below cluster into several conceptual families: supervised approaches (such as Lasso and penalized logistic regression) that select or shrink covariates using their relationship to treatment or outcome; unsupervised linear projections (PCA and random projection) that compress the covariate space without reference to the treatment; locality-aware methods (LPP) that try to preserve neighborhood structure; and nonlinear representation learners (autoencoders and joint causal-representation models) that capture complex covariate geometry at the cost of interpretability. The bias-variance trade-off here mirrors the rest of this chapter. Aggressive compression yields tractable matching but risks discarding information that confounds the treatment-outcome relationship; light compression preserves information but leaves the curse of dimensionality largely intact. As a rule of thumb, supervised sparse methods are sensible defaults when interpretability matters, while nonlinear methods earn their keep only when covariate relationships are genuinely non-linear and the sample is large.

Each method has distinct assumptions and use cases:

-   **Lasso Regression (Least Absolute Shrinkage and Selection Operator)**\
    Lasso imposes an $L_1$ penalty on the regression coefficients, effectively shrinking some coefficients to zero. This property is particularly useful in high-dimensional settings, as it performs both regularization and variable selection.\
    Lasso can be used to identify a smaller set of covariates that are most predictive of treatment assignment or outcome, which can then be used in matching procedures [@gordon2019comparison].

-   **Penalized Logistic Regression**\
    When the treatment is binary, penalized logistic regression (e.g., using an $L_1$ or $L_2$ penalty) can be employed to estimate propensity scores while avoiding overfitting in high-dimensional spaces. These penalized models provide more stable estimates of the propensity score, which is essential for reliable matching [@eckles2021bias].

-   **Principal Component Analysis (PCA)**\
    PCA is an unsupervised linear transformation that projects the original features into a lower-dimensional space by retaining the directions of maximum variance. While PCA does not consider treatment assignment directly, it is effective for denoising and compressing data, particularly when covariates are highly correlated.\
    The principal components can then be used as inputs to standard matching methods.

-   **Locality Preserving Projections (LPP)**\
    LPP is a linear dimensionality reduction technique that, unlike PCA, preserves local neighborhood structures. It constructs a similarity graph and projects data into a lower-dimensional space such that nearby points remain close. This locality-preserving property is beneficial for matching, as it helps maintain the integrity of local relationships within the data [@li2016matching].

-   **Random Projection**\
    Random projection reduces dimensionality by projecting data onto a lower-dimensional subspace using a random matrix. It is computationally efficient and has theoretical guarantees (e.g., Johnson--Lindenstrauss lemma) that distances between points are approximately preserved. This makes it a viable option for extremely high-dimensional data where exact structure preservation is less critical.

-   **Autoencoders**\
    Autoencoders are neural network architectures designed to learn efficient, nonlinear representations (encodings) of data. An autoencoder consists of an encoder that compresses the input and a decoder that attempts to reconstruct the original input from this compressed representation.\
    Autoencoders are particularly effective in capturing complex, nonlinear relationships among features, which may be missed by linear methods like PCA. The latent representation obtained from the encoder can then be used for matching [@ramachandra2018deep].

#### 공동 차원 축소와 분포 균형

Rather than performing dimensionality reduction and matching as separate steps, an emerging strategy is to jointly learn representations that simultaneously:

1.  Reduce dimensionality, and
2.  Balance the covariate distributions between treated and control groups.

This is exemplified by *representation learning for causal inference* approaches. One such method, proposed by @yao2018representation, integrates neural networks with matching objectives to learn latent representations of covariates that are both low-dimensional and balanced. These representations are optimized such that the distributions of treated and control units in the latent space are similar (e.g., using maximum mean discrepancy or other balancing metrics), thereby improving the quality of matches and robustness of treatment effect estimates.

#### 접근법 요약

+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| Method                        | Type            | Supervised? | Key Feature                                     | Reference                |
+===============================+=================+=============+=================================================+==========================+
| Lasso                         | Linear, sparse  | Yes         | Variable selection via $L_1$ penalty            | [@gordon2019comparison]  |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| Penalized logistic regression | Linear          | Yes         | Regularized propensity score estimation         | [@eckles2021bias]        |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| PCA                           | Linear          | No          | Projects onto directions of maximal variance    |                          |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| LPP                           | Linear          | No          | Preserves local neighborhood structure          | [@li2016matching]        |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| Random projection             | Linear (random) | No          | Fast and preserves pairwise distances           |                          |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| Autoencoders                  | Nonlinear       | Yes         | Learns nonlinear representations                | [@ramachandra2018deep]   |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+
| Joint representation learning | Nonlinear       | Yes         | Learns balanced low-dimensional representations | [@yao2018representation] |
+-------------------------------+-----------------+-------------+-------------------------------------------------+--------------------------+

#### 실무 고려사항

-   Model selection and validation: When applying dimensionality reduction prior to matching, one must ensure that the reduced representation still contains sufficient information for confounding adjustment. Cross-validation and balance metrics (e.g., standardized mean differences) should be used to assess the adequacy of the transformation.

-   Interpretability: While methods like PCA or autoencoders can be effective, they may obscure the interpretability of matches since the transformed features may not correspond to original covariates. Sparse methods like Lasso retain interpretability by selecting original covariates.

-   Computational efficiency: Techniques such as random projection or penalized regression are computationally efficient and scalable to large datasets, while autoencoders and joint learning approaches may require more extensive training and hyperparameter tuning.

------------------------------------------------------------------------

### 다중 처치의 매칭

In many applied settings, researchers face the challenge of estimating causal effects for **more than two treatment levels**. For example, a marketing campaign may have three variants (e.g., control, light exposure, heavy exposure), or a policy evaluation may involve multiple interventions. Standard binary treatment matching methods fall short in these cases, necessitating methodological extensions.

Suppose a dataset includes $T$ distinct treatment groups: $\mathcal{T} = \{0, 1, \ldots, T-1\}$. Here, $T = 0$ typically denotes the control group, and $T \in \{1, \ldots, T-1\}$ are active treatments. The goal is to estimate the [Average Treatment Effect] or **Pairwise Treatment Effects**, such as $\text{ATE}_{j,k} = \mathbb{E}[Y(j) - Y(k)]$ for any $j, k \in \mathcal{T}$.

Key challenges in this setting include:

-   Ensuring common support across multiple groups
-   Adjusting for confounders in a balanced way across all treatment pairs
-   Managing covariate imbalance and dimensionality as the number of treatments increases

------------------------------------------------------------------------

#### 다중 처치의 매칭 접근법

Several strategies exist for matching in the presence of multiple treatments, and they differ in how they handle the explosion of pairwise comparisons. Generalized propensity score methods extend the binary scalar logic to a vector of treatment probabilities, which is parsimonious but more demanding on the [overlap](#sec-overlap-positivity-assumption) assumption: every treatment level must have a positive probability of being received at every covariate value. Pairwise matching ignores the multinomial structure and runs a sequence of binary matches, which is simple to implement but can yield matched samples that are inconsistent across pairs. Direct multivariate matching using the full GPS vector preserves the joint structure but is harder to balance. The right choice depends on the number of treatment levels, the available sample within each, and whether the estimand of interest is a single contrast or the full set of pairwise effects.

1.  **Generalized Propensity Scores (GPS)**

The generalized propensity score is defined as the conditional probability of receiving each treatment level given covariates:

$$
e_t(X) = \mathbb{P}(T = t \mid X), \quad \text{for } t = 0, 1, ..., T-1
$$

Estimation is typically done using **multinomial logistic regression**. Once GPS scores are estimated, matching can proceed via:

-   One-vs-all: For each treatment level, match treated units to all others combined.
-   Pairwise matching: Conduct separate pairwise comparisons between all treatment levels.
-   Full matching: Attempt to construct a global matched sample covering all treatments, using the GPS vector.

[@mccaffrey2013tutorial] provides a comprehensive overview of GPS-based methods, including their implementation in the `twang` package.

2.  **Covariate Balancing Propensity Scores (CBPS)** for Multiple Treatments

[@lopez2017estimation] extend CBPS to the multinomial case. Rather than merely estimating GPS, CBPS directly optimizes covariate balance across treatment groups while estimating GPS parameters. This dual-objective estimation leads to more robust causal estimates in finite samples.

3.  **Kernel or Distance-Based Matching on Multinomial Scores**

Instead of reducing GPS to scalar scores, matching can be performed using the full vector of propensity scores, using distance metrics such as Euclidean or Mahalanobis distance in the GPS space. This approach aligns with @zhao2021propensity, who also consider **continuous treatments** using **Generalized Propensity Score Density Estimation**.

------------------------------------------------------------------------

#### MatchIt 및 대안을 사용한 다중 처치 매칭

While the `MatchIt` package in R was originally developed with binary treatment settings in mind, it can be adapted to handle multiple treatment groups through pairwise matching and careful design. However, this approach requires manual data preparation and a clear understanding of the causal estimands of interest, such as the ATT, ATC, or ATE.

1.  Pairwise Matching with a Shared Control Group

To estimate the effect of multiple treatments compared to a shared control group, one straightforward method is to perform separate pairwise matchings between the control group and each treatment group:

```{r}
# Load required libraries
library(MatchIt)
library(cobalt)  # For balance checking

# Example dataset: Simulated
set.seed(123)
n <- 400
df <- data.frame(
  treat = factor(sample(c("control", "A", "B", "C"), n, replace = TRUE)),
  cov1 = rnorm(n),
  cov2 = runif(n),
  cov3 = rbinom(n, 1, 0.5),
  outcome = rnorm(n)
)

# Define treatment levels
treatment_levels <- c("A", "B", "C")
control_level <- "control"

# Initialize weight column
df$match.weights <- 0

# Perform pairwise matching of each treatment group vs control
for (treat in treatment_levels) {
  # Subset to control and current treatment
  subset_df <- df[df$treat %in% c(treat, control_level), ]
  
  # Create a binary treatment variable: 1 for current treatment, 0 for control
  subset_df$treat_binary <- as.numeric(subset_df$treat == treat)

  # Run matching
  m.out <- matchit(treat_binary ~ cov1 + cov2 + cov3,
                   data = subset_df, method = "nearest")

  # Assign weights back to the original dataset
  matched_units <- names(m.out$weights[m.out$weights > 0])
  df[matched_units, "match.weights"] <- m.out$weights[matched_units]
}

# Check covariate balance
bal.tab(treat ~ cov1 + cov2 + cov3, data = df,
        weights = df$match.weights, method = "matching")

# Estimate treatment effects using weighted regression
model <- glm(outcome ~ relevel(treat, ref = "control"),
             data = df[df$match.weights > 0, ],
             weights = match.weights)

summary(model)
```

This approach estimates the **ATT for each treatment group** relative to the control. That is, for each treated group (e.g., A, B, or C), we ask: *what would the outcome have been if those who received treatment had instead received the control?*

2.  Estimating the ATC Using `MatchIt`

In some cases, we may wish to estimate the ATC, i.e., what would have happened to the control group if they had received each of the treatments. To do this, we match control units to each treated group, reversing the focal population. Practically, this means keeping the control group intact and separately matching each treatment group to it.

```{r}
# Load required libraries
library(MatchIt)
library(cobalt)

# Simulate example data
set.seed(456)
n <- 400
df <- data.frame(
  treat = factor(sample(c("control", "A", "B", "C"), n, replace = TRUE)),
  cov1 = rnorm(n),
  cov2 = runif(n),
  cov3 = rbinom(n, 1, 0.5),
  outcome = rnorm(n)
)

# Define treatment levels
treatment_levels <- c("A", "B", "C")
control_level <- "control"

# Initialize weight column
df$match.weights <- 0

# Estimate ATC by matching treated units to the control group
for (treat in treatment_levels) {
  # Subset to current treatment and control
  subset_df <- df[df$treat %in% c(treat, control_level), ]
  
  # Binary treatment variable: 0 for treatment group, 1 for control
  # This reverses the focus, targeting the control as the treated group
  subset_df$treat_binary <- as.numeric(subset_df$treat == control_level)

  # Perform matching
  m.out <- matchit(treat_binary ~ cov1 + cov2 + cov3,
                   data = subset_df, method = "nearest")
  
  # Extract matched unit IDs
  matched_ids <- names(m.out$weights[m.out$weights > 0])
  
  # Assign weights back to original dataset
  df[matched_ids, "match.weights"] <- m.out$weights[matched_ids]
}

# Check balance (optional)
bal.tab(treat ~ cov1 + cov2 + cov3, data = df,
        weights = df$match.weights, method = "matching")

# Estimate ATC via weighted regression
model <- glm(outcome ~ relevel(treat, ref = "control"),
             data = df[df$match.weights > 0, ],
             weights = match.weights)

summary(model)

```

3.  Regression and Effect Estimation

After constructing the matched dataset, one can use regression on the matched sample to estimate the treatment effects:

```{r}
# Subset the data first
matched_data <- df[df$match.weights > 0, ]

# Then fit the weighted regression using the weights from the subset
model <- glm(outcome ~ relevel(treat, ref = "control"),
             data = matched_data,
             weights = matched_data$match.weights)

summary(model)
```

Caveats and Limitations

-   Matching one treatment group at a time does not preserve global covariate balance across all treatment groups, only pairwise balance with the control.

-   Overlap (common support) assumptions should be verified individually for each pairwise comparison.

-   Matched samples may vary for each comparison, complicating aggregate inference or joint hypothesis testing.

------------------------------------------------------------------------

#### 대안: `WeightIt`을 사용한 다중 처치 가중치 부여

Matching can be cumbersome in multiple-treatment settings. Weighting approaches offer a more seamless framework, especially when estimating ATE or ATC across all treatment levels simultaneously. The `WeightIt` package extends propensity score weighting to multi-treatment scenarios with strong support for diagnostics and flexible estimation.

```{r}
# Load required libraries
library(WeightIt)
library(cobalt)

# Simulate example data
set.seed(789)
n <- 400
df <- data.frame(
  treat = factor(sample(c("control", "A", "B", "C"), n, replace = TRUE)),
  cov1 = rnorm(n),
  cov2 = runif(n),
  cov3 = rbinom(n, 1, 0.5),
  outcome = rnorm(n)
)

# Estimate weights for ATE across all treatment levels using multinomial logistic regression
w.out <- weightit(
  treat ~ cov1 + cov2 + cov3,
  data = df,
  estimand = "ATE",
  method = "glm"  # multinomial model when treat is a factor with >2 levels
)

# Check covariate balance
bal.tab(w.out)

bal.tab(w.out, which.treat = c("A", "B", "C"))


# Estimate treatment effects with robust SEs
library(jtools)
model <- glm(outcome ~ relevel(treat, ref = "control"),
             data = df,
             weights = w.out$weights)
summ(model, robust = "HC1")
```

Additional Notes:

-   To estimate the ATT for a specific treatment (e.g., treatment A), set `focal = "A"` and `estimand = "ATT"` in `weightit()`.

-   Setting `method = "gbm"` in `WeightIt` will use boosted models, equivalent to the `twang` package.

-   Weighting approaches avoid sample size loss from discarding unmatched units and provide smoother covariate balance optimization.

#### 요약: 다중 처치 환경에서의 매칭과 가중치 부여

+------------+------------------------------------+---------------------+----------------------------------------+---------------------------------------------------------+
| Method     | Approach                           | Estimands Supported | Pros                                   | Cons                                                    |
+============+====================================+=====================+========================================+=========================================================+
| `MatchIt`  | Pairwise matching (manual)         | ATT, ATC            | Intuitive; transparent matched pairs   | Manual; repeated steps; limited to pairwise comparisons |
+------------+------------------------------------+---------------------+----------------------------------------+---------------------------------------------------------+
| `WeightIt` | Simultaneous weighting             | ATT, ATC, ATE       | Flexible; single model; scalable       | Requires model diagnostics                              |
+------------+------------------------------------+---------------------+----------------------------------------+---------------------------------------------------------+
| `twang`    | Boosted weighting for GPS          | ATT, ATE            | Automatic tuning; longitudinal support | More complex interface                                  |
+------------+------------------------------------+---------------------+----------------------------------------+---------------------------------------------------------+
| `CBPS`     | Balance-constrained GPS estimation | ATT, ATE            | Direct balance optimization            | May require customization                               |
+------------+------------------------------------+---------------------+----------------------------------------+---------------------------------------------------------+

In general, weighting is more scalable and coherent for estimating causal effects in multi-treatment contexts, while matching can be useful when interpretability and transparency of individual pairings is important.

------------------------------------------------------------------------

### 다수준 처치의 매칭

Some treatments are not just categorical, but **ordinal** (multi-level), such as "low", "medium", and "high" dosage levels. These differ from multinomial treatments in that the levels have a **natural order**. Incorporating this structure can improve both estimation efficiency and interpretability.

#### 성향점수 추정

In this context, researchers often use **ordinal logistic regression** to estimate the probability of being in each treatment level:

$$
\text{logit}\left(\mathbb{P}(T \leq t \mid X)\right) = \alpha_t - X^\top \beta
$$

This model imposes the **proportional odds assumption**, where the log-odds are linear in covariates but share the same coefficients across cutoffs.

#### 매칭 전략

@yang2016propensity provide a framework for matching with ordinal treatments, arguing that matching on the **generalized propensity score** derived from an ordinal model helps preserve the order and improves balance. Key considerations include:

-   Matching on estimated ordinal probabilities or cumulative logits
-   Using Mahalanobis distance in the latent score space
-   Implementing stratification or subclassification based on predicted scores

A custom package, [`shuyang1987/multilevelMatching`](https://github.com/shuyang1987/multilevelMatching), provides tools for this kind of analysis. The package includes:

-   Functions to estimate ordinal GPS
-   Matching and subclassification routines
-   Diagnostics to evaluate covariate balance

------------------------------------------------------------------------

### 연속형 처치의 매칭 {#sec-gps-continuous}

The natural endpoint of generalizing the propensity score from binary to multinomial to ordinal treatments is the case where the treatment is fully continuous: a dose $D \in \mathcal{D} \subset \mathbb{R}$, not a level of an ordered factor. The substantive question shifts accordingly. The deliverable is no longer a single contrast or a small set of pairwise effects but an entire **dose-response function** $\mu(d) = \mathbb{E}[Y(d)]$ giving the average potential outcome at each dose. Reporting a single number when the data support a curve is a missed opportunity at best and an averaging-away of heterogeneity at worst.

The cross-sectional framework developed below complements the [continuous-treatment difference-in-differences](#sec-continuous-did) machinery in the DiD chapter. The two address the same substantive question (how does the average effect scale with the dose?) but make different identifying bets: the panel-data DiD relies on parallel trends across doses, while the matching framework here relies on a conditional-ignorability extension of the binary-treatment unconfoundedness assumption. In applications where both are feasible, reporting both is informative; in most applications, the data structure dictates which is available.

#### 연속형 처치의 일반화 성향점수

@hirano2004propensity introduced the generalized propensity score for continuous treatments and showed that the central balancing property of the binary propensity score carries over with only a notational change. The GPS is now defined as the conditional density of the treatment given the covariates, evaluated at the treatment actually received:

$$
R = r(D, X), \quad \text{where} \quad r(d, x) = f_{D \mid X}(d \mid x).
$$

The intuition is the same as in the binary case. Units that, conditional on covariates, were equally likely to receive a given dose are comparable at that dose. The GPS reduces all covariate information relevant for the assignment mechanism to a single scalar that is sufficient for balancing.

The framework relies on a **weak unconfoundedness** assumption, which is the natural lift of the binary-treatment ignorability condition:

$$
Y(d) \perp\!\!\!\perp D \mid X, \quad \forall d \in \mathcal{D}.
$$

In words, within strata defined by the covariates, the dose received is independent of the potential outcomes at every dose. The assumption is "weak" only relative to a stronger joint independence; in practice it is no more or less testable than the binary unconfoundedness assumption it generalizes, and it relies on the same substantive defense: that the analyst has measured everything that jointly drives selection into dose and the outcome.

@hirano2004propensity prove the central balancing result: within strata of the GPS $R$, the dose $D$ is independent of the covariates $X$. The implication is the same as in the binary case: conditioning on $R$ in an outcome model is sufficient to remove the confounding bias from the covariates, even though $R$ is a one-dimensional summary of a potentially high-dimensional $X$. The dose-response function is then identified by

$$
\mu(d) = \mathbb{E}\left[\,\mathbb{E}[Y \mid D = d, R = r(d, X)]\,\right],
$$

where the outer expectation is over the marginal distribution of the covariates (equivalently, over the marginal distribution of $r(d, X)$ at the chosen $d$).

@imai2004causal develop a complementary framework that subsumes binary, multinomial, ordinal, and continuous treatments in a single notation by treating the entire treatment-assignment density as the object of interest. The two frameworks coincide in the continuous case and differ mainly in notational emphasis; the @hirano2004propensity formulation is more common in applied work.

#### 실제 추정량

Three estimation strategies dominate applied work, ordered by how much functional-form discipline they impose.

The first is the **parametric two-step estimator** of @hirano2004propensity. The GPS is modeled with a parametric density (typically a normal or log-normal conditional on covariates, with the parameters estimated by maximum likelihood). The outcome model is then a flexible parametric regression of $Y$ on a polynomial in $(D, R)$, often quadratic in each. The dose-response curve is recovered by averaging the fitted outcome model over the empirical distribution of the covariates, holding $D$ fixed at each value in a grid. This estimator is computationally cheap and has the asymptotic guarantees one expects from a well-specified parametric model; its weakness is that it inherits the standard parametric-model risk of model misspecification, which in the continuous case is harder to diagnose than in the binary case.

The second is the **covariate-balancing GPS** of @lopez2017estimation and related extensions of CBPS to continuous treatments. The conditional-density model is fit not by likelihood but by minimizing an explicit measure of imbalance between the treated and control distributions across the dose range. The resulting weights produce a pseudo-population in which the dose is approximately independent of the covariates, which lets the analyst estimate the dose-response with a simple weighted regression. The advantage is that covariate balance, the object the analyst actually cares about, is the target of optimization rather than a hoped-for consequence of correct density specification.

The third is the **nonparametric doubly robust estimator** of @kennedy2017nonparametric. The dose-response is estimated by a two-step procedure that combines a nonparametric outcome model with a nonparametric density model for the GPS, with cross-fitting to deliver valid inference under flexible (e.g., machine-learning-based) first-stage estimators. The estimator has the doubly-robust property that consistency holds if either the outcome model or the density model is correctly specified, and the nonparametric flexibility addresses the most common failure mode of the parametric two-step. The cost is larger sample-size requirements and computational expense.

The applied literature provides useful templates. @kluve2012evaluating apply the @hirano2004propensity framework to a continuous-dose training programme in Germany, recovering an inverse-U dose-response in which moderate training duration is most effective. @flores2012estimating use the GPS framework to estimate dose-response curves of the Job Corps programme on labor-market outcomes, with the dose measured as the length of exposure to instruction. Both papers serve as detailed methodological worked examples in addition to their substantive contributions.

#### 예제: 시연을 먼저, 실제 적용을 다음으로

We work through the GPS dose-response estimator in two passes. The first pass implements the @hirano2004propensity two-step procedure from scratch in base R, so that the mechanics of the GPS, the balancing property, and the dose-response averaging are all visible. The second pass calls the published `CausalGPS` package on the same data.

##### Demonstration: Hirano-Imbens GPS from scratch

```{r gps-cont-sim, message=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tidyr)
})

set.seed(20240515)

# Generate synthetic data: continuous treatment D depends on three
# confounders X1, X2, X3. The true dose-response is quadratic in D
# with a maximum near D = 2 (an inverse-U pattern, common in training
# programmes).
n <- 2000
X1 <- rnorm(n)
X2 <- rnorm(n)
X3 <- rbinom(n, 1, 0.4)

# Treatment-assignment model: D ~ Normal( mu(X), sigma^2 ).
mu_D    <- 1.5 + 0.6 * X1 - 0.4 * X2 + 0.5 * X3
sigma_D <- 1.0
D       <- rnorm(n, mu_D, sigma_D)

# Outcome model: quadratic in D, with confounding via X.
Y <- 0.3 + 1.2 * D - 0.3 * D^2 + 0.5 * X1 + 0.2 * X2 - 0.1 * X3 +
     rnorm(n, sd = 0.5)

dat <- data.frame(Y, D, X1, X2, X3)
head(dat)
```

The first step of the @hirano2004propensity procedure is to estimate the GPS, the conditional density of $D$ given $X$, evaluated at each unit's actual dose. We assume a normal density and fit $\mu(X)$ by OLS:

```{r gps-cont-estimate}
# Step 1: fit the conditional mean of D given X.
gps_mean_model <- lm(D ~ X1 + X2 + X3, data = dat)
mu_hat         <- predict(gps_mean_model)
sigma_hat      <- sd(residuals(gps_mean_model))

# GPS = f_{D|X}(D | X), evaluated at the actual treatment.
dat$R <- dnorm(dat$D, mean = mu_hat, sd = sigma_hat)
summary(dat$R)
```

The `summary(dat$R)` output reports the distribution of the estimated GPS across the sample: it should be a positive, smooth distribution with no point masses, since $D$ given $X$ is modelled as continuous. Extremely small GPS values flag units with implausible dose-covariate combinations (the continuous-treatment analogue of extreme propensity scores in the binary case).

To diagnose the balancing property, we stratify on the estimated GPS and check that, within strata, the covariates are no longer strongly associated with the dose:

```{r gps-cont-balance}
dat$R_quint <- ntile(dat$R, 5)

balance_check <- dat %>%
  group_by(R_quint) %>%
  summarise(
    cor_D_X1 = cor(D, X1),
    cor_D_X2 = cor(D, X2),
    cor_D_X3 = cor(D, X3),
    n        = n()
  )

# For comparison, the raw correlations before adjustment:
raw_cor <- data.frame(
  R_quint  = "RAW (no adjustment)",
  cor_D_X1 = cor(dat$D, dat$X1),
  cor_D_X2 = cor(dat$D, dat$X2),
  cor_D_X3 = cor(dat$D, dat$X3),
  n        = nrow(dat)
)

bind_rows(raw_cor, mutate(balance_check, R_quint = as.character(R_quint)))
```

Reading the table: the top row shows the unadjusted correlations between $D$ and each covariate (the confounding we are trying to remove). The next five rows show the within-GPS-stratum correlations. If the balancing property holds, the stratum correlations should be much closer to zero than the raw correlations. The pattern is exactly what we expect from theory: conditioning on a one-dimensional summary of the assignment mechanism is sufficient to balance the covariates that drove the assignment.

The second step of the procedure is the outcome model. We fit a flexible regression of $Y$ on a polynomial in $(D, R)$:

```{r gps-cont-outcome}
outcome_model <- lm(Y ~ D + I(D^2) + R + I(R^2) + D:R, data = dat)
summary(outcome_model)$coefficients
```

The coefficient table reports the fitted parametric outcome model; individual coefficients are not the substantive object of interest, but the fitted surface is. The dose-response curve is recovered by averaging the fitted outcome over the empirical covariate distribution, holding $D$ fixed at each value in a grid:

```{r gps-cont-dose-response, fig.cap="Estimated dose-response curve from the Hirano-Imbens GPS two-step. Solid line is the fitted average potential outcome at each dose; dashed red is the true quadratic dose-response used in the simulation. The estimator recovers the inverse-U shape and the location of the maximum.", fig.width=6, fig.height=4}
d_grid <- seq(min(dat$D), max(dat$D), length.out = 100)

# For each grid value of D, predict the outcome at every unit holding D
# fixed at the grid value (and recomputing the GPS at the counterfactual
# dose), then average over the empirical covariate distribution. The mean
# of the predictions IS the dose-response value at that dose.
dose_response <- sapply(d_grid, function(d_val) {
  r_at_d <- dnorm(d_val, mean = mu_hat, sd = sigma_hat)
  preds  <- predict(outcome_model,
                    newdata = data.frame(D = d_val, R = r_at_d))
  mean(preds)
})

curve_df <- data.frame(
  D      = d_grid,
  Y_hat  = dose_response,
  Y_true = 0.3 + 1.2 * d_grid - 0.3 * d_grid^2
)

ggplot(curve_df, aes(x = D)) +
  geom_line(aes(y = Y_hat), linewidth = 1) +
  geom_line(aes(y = Y_true), linetype = "dashed", color = "red") +
  labs(x = "Dose D",
       y = "E[Y(d)]: average potential outcome at dose d",
       title = "GPS dose-response (Hirano-Imbens two-step, by hand)",
       subtitle = "Solid: fitted dose-response. Dashed red: true curve.") +
  theme_minimal()
```

Figure \@ref(fig:gps-cont-dose-response) compares the estimated dose-response to the true curve. The estimator recovers both the shape (inverse-U) and the approximate location of the maximum, demonstrating that the GPS-stratification logic does the work it is meant to do. In a real application, of course, the true curve is unknown; the analyst reports only the solid line, typically with a bootstrap confidence band.

##### Production: `CausalGPS` package

For production work we recommend the `CausalGPS` package on CRAN by Khoshnevis, Wu, and Braun, which implements both matching and weighting versions of the GPS-based dose-response analysis, with support for flexible (machine-learning-based) density estimation and a built-in nonparametric estimator for the exposure-response function. The pipeline has four steps: estimate the GPS, compute matching or weighting counter-weights, generate the pseudo-population, and estimate the dose-response curve.

```{r causalgps-pipeline, eval=FALSE}
# install.packages("CausalGPS")  # CRAN, Khoshnevis, Wu, Braun
library(CausalGPS)

set.seed(20240515)

# Synthetic data: continuous exposure w, six confounders, outcome Y
m_d <- generate_syn_data(sample_size = 1000)
head(m_d)

# Step 1: Estimate the generalized propensity score.
# Here a gradient-boosted (XGBoost) density model is used, with a normal
# density assumption for the conditional distribution of the exposure.
gps_obj <- estimate_gps(
  .data       = m_d,
  .formula    = w ~ I(cf1^2) + cf2 + I(cf3^2) + cf4 + cf5 + cf6,
  sl_lib      = c("m_xgboost"),
  gps_density = "normal"
)

# Step 2: Compute matching weights. The matching is performed in the
# (exposure, GPS) space using L1 distance; delta_n controls the
# exposure-side bandwidth, scale balances exposure vs. GPS proximity.
cw_object <- compute_counter_weight(
  gps_obj      = gps_obj,
  ci_appr      = "matching",
  delta_n      = 0.1,
  dist_measure = "l1",
  scale        = 0.5
)

# Step 3: Build the pseudo-population (each row carries a counter weight
# representing how many units like it would appear in the rebalanced
# population at the dose it actually received).
pseudo_pop <- generate_pseudo_pop(
  .data               = m_d,
  cw_obj              = cw_object,
  covariate_col_names = c("cf1", "cf2", "cf3", "cf4", "cf5", "cf6")
)

# Step 4: Estimate the exposure-response curve at a grid of dose values.
erf_obj <- estimate_erf(
  .data           = pseudo_pop$.data,
  .formula        = Y ~ w,
  weights_col_name = "counter_weight",
  model_type      = "nonparametric",
  w_vals          = seq(2, 20, 0.5),
  kernel_appr     = "kernsmooth"
)

plot(erf_obj)
```

The figure returned by `plot(erf_obj)` is the analogue of the dose-response figure produced in the from-scratch demonstration above (and of `contdid::ggcont_did()` in the panel-data setting): a curve $\hat{\mu}(d) = \hat{\mathbb{E}}[Y(d)]$ at each grid value of the dose, with a confidence band reflecting the sampling uncertainty in both the GPS estimation step and the outcome estimation step. The right way to read it is to look for the dose range over which the band excludes any substantively interesting null (zero, or a policy benchmark) and to report the shape of the curve rather than any single point. The `summary(erf_obj)` companion returns a table with the grid values of the dose, the point estimate, and the band limits, suitable for tabular reporting alongside the figure.

Two further outputs are useful in production. First, `gnomeplot()` (or the package's `check_covar_balance()` function) returns a covariate-balance diagnostic that plots the absolute standardized mean differences of each covariate against the dose, before and after rebalancing; the right interpretation is that every covariate's adjusted curve should hover near zero across the entire dose range, with deviations flagging doses at which the GPS model failed to balance. Second, `pseudo_pop$adjusted_corr_results` reports the absolute weighted correlation between each covariate and the dose in the pseudo-population; values below 0.1 are conventional for asserting that balance has been achieved.

For weighting (rather than matching) the only change is the `ci_appr` argument:

```{r causalgps-weighting, eval=FALSE}
cw_weight <- compute_counter_weight(
  gps_obj = gps_obj,
  ci_appr = "weighting"
)
```

Weighting tends to be more efficient when the dose distribution is well populated everywhere in the support of interest; matching tends to be more robust when the dose distribution has thin regions, because it makes the loss of effective sample size explicit rather than hiding it in extreme weights.

#### 진단

The diagnostic that matters most in continuous-treatment GPS analysis is **covariate balance as a function of the dose**, not balance averaged over the entire pseudo-population. The right object to plot is the absolute standardized mean difference of each covariate between units near a target dose and units away from it, evaluated at several doses spanning the support. A pseudo-population that is balanced on average but unbalanced at the doses where the dose-response curve has its most interesting features (typically the tails) is not balanced where the analyst needs it to be. `CausalGPS` provides built-in balance diagnostics that report this dose-stratified balance directly.

A second diagnostic is the **effective sample size after rebalancing**, both globally and at each dose. A pseudo-population that achieves balance only by upweighting a tiny subset of the data is fragile; the dose-response confidence intervals will be wide and sensitive to small perturbations.

The third diagnostic is a **common-support check**: identify the range of doses over which units of every covariate profile have positive (or at least non-negligible) probability of being observed. The dose-response is identified only over this range; extrapolating beyond it is a statement about the model, not about the data.

------------------------------------------------------------------------

### 반복 처치의 매칭(시간가변 처치)

In longitudinal studies, treatments are often administered at multiple time points. Examples include:

-   A patient receiving a drug in multiple doses over weeks
-   A user receiving marketing emails over several days

This setting raises unique challenges, such as:

-   Time-varying confounding: Covariates affected by prior treatment may influence future treatment and outcomes
-   Cumulative dose effects: Treatment assignment is no longer a one-time event

#### 주변 구조 모형

The most popular framework for analyzing repeated treatments is the **Marginal Structural Model (MSMs)**, which estimates causal effects by weighting each observation using **Inverse Probability of Treatment Weights (IPTW)**.

Let $A_t$ be the treatment at time $t$, and $X_t$ be time-varying covariates. The IPTW for a trajectory is:

$$
w_i = \prod_{t=1}^{T} \frac{1}{\mathbb{P}(A_{it} \mid \bar{A}_{i,t-1}, \bar{X}_{i,t})}
$$

Weights are estimated using logistic regression models at each time point. The outcome model then regresses $Y$ on $\bar{A}_T$ using the weights.

The [`twang`](https://cran.r-project.org/web/packages/twang/vignettes/iptw.pdf) package provides tools for:

-   Estimating time-varying propensity scores
-   Computing IPTWs
-   Fitting marginal structural models
-   Checking covariate balance over time

Practical Notes

-   Stabilized weights help reduce variance
-   Trimming or truncating extreme weights is often necessary to maintain robust inference
-   Dynamic treatment regimes may require further generalizations such as structural nested mean models.

------------------------------------------------------------------------

## 비관측변수에 의한 선택 {#sec-selection-on-unobservables}

While randomized experiments help eliminate bias from unobserved factors, observational data often leave us vulnerable to [selection on unobservables](#sec-selection-on-unobservables), a scenario where omitted variables jointly affect both treatment assignment and outcomes. Unlike [selection on observables](#sec-selection-on-observables), these unobserved confounders cannot be directly controlled by regression or matching techniques.

One popular strategy to mitigate bias from observable variables is [matching methods](#sec-matching-methods). These methods are widely used to estimate causal effects under the assumption of **selection on observables** and offer two primary advantages:

-   They reduce reliance on functional form assumptions (unlike parametric regression).
-   They rely on the assumption that all covariates influencing treatment assignment are observed.

However, this key assumption is often unrealistic. When relevant covariates are unmeasured, **bias from unobservables remains a threat**. To address this concern, researchers turn to **Rosenbaum Bounds**, a sensitivity analysis framework that quantifies how strong hidden bias must be to explain away the observed treatment effect.

Several econometric methods have been developed to test the robustness of causal estimates against such hidden bias. Below are key strategies that researchers commonly employ:

1.  [Rosenbaum Bounds](#sec-rosenbaum-bounds)\
    This approach assesses the sensitivity of treatment effects to potential hidden bias by bounding the treatment effect under varying assumptions about the strength of unobserved confounding. It is particularly useful when treatment assignment is believed to be *conditionally ignorable*, but this assumption may be imperfect.

2.  [Endogenous Sample Selection] (Heckman-style correction)\
    When sample selection is non-random and correlated with unobserved variables influencing the outcome, the Heckman selection model provides a correction. It introduces the inverse Mills ratio $\lambda$ as a control function. A statistically significant $\lambda$ indicates that unobserved factors influencing selection also affect the outcome, pointing to endogenous selection.

    -   Interpretation tip: A significant $\lambda$ suggests selection bias due to unobservables, and the Heckman model adjusts the outcome equation accordingly.
    -   This method is frequently applied in labor economics and marketing analytics (e.g., correcting for self-selection in surveys or consumer choice modeling).

3.  [Relative Correlation Restrictions](#sec-relative-correlation-restrictions)\
    This method imposes assumptions about the correlation between observed and unobserved variables. For instance, by assuming that the selection on unobservables is not stronger than the selection on observables (as in @altonji2005selection), one can identify bounds on causal effects. These approaches often rely on auxiliary assumptions or symmetry restrictions to enable partial identification.

4.  [Coefficient-Stability Bounds](#sec-coefficient-stability-bounds)\
    This technique evaluates how treatment effect estimates change with the inclusion of additional control variables. If the coefficient on the treatment variable remains stable as more covariates are added, it suggests robustness to omitted variable bias.

    -   This idea underlies methods like Oster's $\delta$ method [@oster2019unobservable], which quantifies how much selection on unobservables would be required to nullify the observed effect, assuming proportional selection between observables and unobservables.

In applied business research, such as customer retention, credit scoring, or pricing analytics, selection on unobservables is often unavoidable. These methods do not *eliminate* the problem but provide frameworks to quantify and interpret the risk of hidden bias. The choice among these tools depends on the specific context, data availability, and the plausibility of underlying assumptions.

------------------------------------------------------------------------

### Rosenbaum 경계 {#sec-rosenbaum-bounds}

Rosenbaum Bounds assess the **robustness of treatment effect estimates to hidden bias** by introducing a sensitivity parameter $\Gamma$ (gamma), which captures the potential effect of an unobserved variable on treatment assignment.

-   $\Gamma = 1$: Indicates perfect randomization, where units in a matched pair are equally likely to receive treatment.
-   $\Gamma = 2$: Suggests that one unit in a matched pair could be up to **twice as likely** to receive treatment due to an unmeasured confounder.

Since $\Gamma$ is unknown, sensitivity analysis proceeds by varying $\Gamma$ and examining how the statistical significance and magnitude of the treatment effect respond. This allows us to evaluate how strong an unobserved confounder would need to be to invalidate our findings.

Because the unobserved variable must affect both selection into treatment and outcomes, Rosenbaum Bounds are often described as a **worst-case analysis** [@diprete2004assessing].

------------------------------------------------------------------------

Imagine we match two individuals based on age and education. One receives treatment, the other does not. If there's an unmeasured trait (say, motivation) that makes one of them **twice** as likely to receive treatment, this is equivalent to $\Gamma = 2$.

We then ask: *At what level of* $\Gamma$ does our result lose significance? If that level is very high (e.g., $\Gamma = 3$ or more), we can be more confident in our findings.

------------------------------------------------------------------------

In layman's terms, consider two matched individuals who differ only in an unobserved trait (e.g., motivation). If one of them is twice as likely to receive treatment because of this unobservable, this corresponds to $\Gamma = 2$. We then ask: *Would the treatment effect remain significant under this level of hidden bias?*

If the estimated effect becomes insignificant only at high values of $\Gamma$ (e.g., $\Gamma > 2$), the effect is considered robust. Conversely, if statistical significance vanishes at $\Gamma \approx 1.1$, the effect is highly sensitive to hidden bias.

------------------------------------------------------------------------

#### 기술 요약

-   Rosenbaum Bounds assess sensitivity without requiring knowledge of the **direction** of the unobserved bias.
-   They apply to **matched data**, typically using **Wilcoxon signed-rank tests** or other non-parametric statistics.
-   In the presence of **random treatment assignment**, such non-parametric tests are valid directly.
-   In **observational data**, they are valid under the assumption of **selection on observables**. If that assumption is questionable, Rosenbaum Bounds offer a way to **test its believability** [@rosenbaum2002attributing].

A typical Rosenbaum Bounds analysis presents:

-   The $p$-value for statistical significance at increasing levels of $\Gamma$.
-   The Hodges-Lehmann point estimate (a robust median-based treatment effect estimator).
-   The critical $\Gamma$ value at which the treatment effect becomes insignificant.

These analyses do not provide exact bounds on treatment effect estimates. For that, other approaches such as [Relative Correlation Restrictions](#sec-selection-on-unobservables) are required.

------------------------------------------------------------------------

#### 기술적 프레임워크

Let $\pi_i$ denote the probability that unit $i$ receives treatment. The odds of treatment are:

$$
\text{Odds}_i = \frac{\pi_i}{1 - \pi_i}
$$

Ideally, after matching, if there's no hidden bias, $\pi_i = \pi_j$ for unit $i$ and $j$.

Rosenbaum bounds constrain the **odds ratio** between two matched units $i$ and $j$:

$$
\frac{1}{\Gamma} \le \frac{\text{Odds}_i}{\text{Odds}_j} \le \Gamma
$$

This relationship links the hidden bias directly to selection probabilities.

Suppose treatment assignment follows:

$$
\log\left( \frac{\pi_i}{1 - \pi_i} \right) = \kappa x_i + \gamma u_i
$$

where $u_i$ is an unobserved covariate, and $x_i$ is an observed covariate. The greater the $\gamma$, the stronger the influence of the unobservable.

------------------------------------------------------------------------

#### 민감도 분석 절차

1.  Choose a plausible range of $\Gamma$ (typically from 1 to 2.5).
2.  Assess how the p-value or the magnitude of the treatment effect [@hodges2011estimates] (for more details, see @hollander2013nonparametric) changes with varying $\Gamma$ values.
3.  Employ specific randomization tests based on the type of outcome to establish bounds on inferences.
    -   Report the minimum value of $\Gamma$ at which the treatment effect is nullified (i.e., becomes insignificant). And the literature's rules of thumb is that if $\Gamma > 2$, then we have strong evidence for our treatment effect is robust to large biases [@proserpio2017online].

If treatment is clustered (e.g., by region or school), standard Rosenbaum bounds must be adjusted. See [@hansen2014clustered] for methods that extend sensitivity analysis to clustered settings, analogous to using clustered standard errors in regression.

------------------------------------------------------------------------

#### 마케팅 실증 예제

Table \@ref(tab:rosenbaum-gamma-marketing) reports $\Gamma$ thresholds needed to nullify treatment effects in real-world marketing studies.

Table: (\#tab:rosenbaum-gamma-marketing) Critical Rosenbaum $\Gamma$ values from selected marketing studies.

+---------------------------+-------------------+-----------------------------------------------------------+
| Study                     | Critical $\Gamma$ | Context                                                   |
+===========================+===================+===========================================================+
| [@oestreicher2013content] | 1.5 to 1.8        | User community participation on willingness to pay        |
+---------------------------+-------------------+-----------------------------------------------------------+
| [@sun2013ad]              | 1.5               | Revenue-sharing program and content popularity            |
+---------------------------+-------------------+-----------------------------------------------------------+
| [@manchanda2015social]    | 1.6               | Impact of social referrals on purchase behavior           |
+---------------------------+-------------------+-----------------------------------------------------------+
| [@sudhir2015peter]        | 1.9 to 2.2        | IT adoption effects on productivity                       |
+---------------------------+-------------------+-----------------------------------------------------------+
| [@proserpio2017online]    | 2.0               | Management responses and online reputation                |
+---------------------------+-------------------+-----------------------------------------------------------+
| [@zhang2022makes]         | 1.55              | Verified photos and Airbnb demand                         |
+---------------------------+-------------------+-----------------------------------------------------------+
| [@chae2023paywall]        | 27.0              | Paywall suspensions and future subscriptions (not a typo) |
+---------------------------+-------------------+-----------------------------------------------------------+

Packages

-   `rbounds` [@keele2010overview]

-   `sensitivitymv` [@rosenbaum2015two]

Since we typically assess our estimate sensitivity to unobservables after matching, we first do some matching.

```{r}
library(MatchIt)
library(Matching)
data("lalonde", package = "MatchIt")

# Matching
matched <- matchit(treat ~ age + educ, data = MatchIt::lalonde, method = "nearest")
summary(matched)
matched_data <- match.data(matched)

# Split into groups
treatment_group <- subset(matched_data, treat == 1)
control_group   <- subset(matched_data, treat == 0)

# Load rbounds package
library(rbounds)

# P-value sensitivity
psens_res <- psens(treatment_group$re78,
                   control_group$re78,
                   Gamma = 2,
                   GammaInc = 0.1)
psens_res

# Hodges-Lehmann estimate sensitivity
hlsens_res <- hlsens(treatment_group$re78,
                     control_group$re78,
                     Gamma = 2,
                     GammaInc = 0.1)
hlsens_res
```

For matching with more than one control per treated unit:

```{r}
library(Matching)
library(MatchIt)

n_ratio <- 2
matched <-
    matchit(treat ~ age + educ,
            data = MatchIt::lalonde,
            method = "nearest",
            ratio = n_ratio)
matched_data <- match.data(matched)

mcontrol_res <- rbounds::mcontrol(
    y          = matched_data$re78,
    grp.id     = matched_data$subclass,
    treat.id   = matched_data$treat,
    group.size = n_ratio + 1,
    Gamma      = 2.5,
    GammaInc   = 0.1
)
mcontrol_res

```

Alternative Sensitivity Analysis Tools

`sensitivitymv` is faster than `sensitivitymw`. But `sensitivitymw` can match where matched sets can have differing numbers of controls [@rosenbaum2015two].

```{r}
library(sensitivitymv)
data(lead150)
senmv(lead150, gamma = 2, trim = 2)

library(sensitivitymw)
senmw(lead150, gamma = 2, trim = 2)

```

------------------------------------------------------------------------

### 상대 상관 제한 {#sec-relative-correlation-restrictions}

In many observational studies, researchers are concerned about the impact of unobserved variables that are not included in the regression model. Even after including a comprehensive set of controls, there is always the possibility that omitted variables still bias the estimated treatment effect. **Relative Correlation Restrictions (RCR)** offer a formal way to assess the **potential magnitude of this bias** by comparing it to the observed selection.

------------------------------------------------------------------------

Let's begin with a linear outcome model:

$$
Y_i = X_i \beta + C_i \gamma + \epsilon_i
$$

Where:

-   $Y_i$ is the outcome variable (e.g., revenue, engagement, churn),
-   $X_i$ is the treatment or variable of interest (e.g., paywall suspension, ad exposure),
-   $C_i$ is a set of observed control covariates,
-   $\epsilon_i$ is the error term, which may include unobserved factors.

In standard OLS regression, we assume:

$$
\text{Cov}(X_i, \epsilon_i) = 0
$$

That is, the treatment $X_i$ is uncorrelated with the omitted variables contained in $\epsilon_i$. This assumption is crucial for the **unbiased estimation of** $\beta$, the treatment effect.

However, in observational settings, this assumption may not hold. There may be unobserved confounders that influence both $X_i$ and $Y_i$, biasing the OLS estimate of $\beta$. The Relative Correlation Restrictions framework provides a way to bound the treatment effect under assumptions about the relative strength of selection on unobservables compared to selection on observables.

------------------------------------------------------------------------

The method was originally proposed by @altonji2005selection and later extended by @krauth2016bounding. The core idea is to assume a proportional relationship between the correlation of the treatment with the error term and its correlation with the observed controls:

$$
\text{Corr}(X_i, \epsilon_i) = \lambda \cdot \text{Corr}(X_i, C_i \gamma)
$$

Here, $\lambda$ represents the relative strength of selection on unobservables compared to observables.

-   If $\lambda = 0$, then we return to the standard OLS assumption: no unobserved selection.
-   If $\lambda = 1$, we assume that selection on unobservables is as strong as selection on observables.
-   If $\lambda > 1$, then unobserved selection is assumed to be stronger than observable selection.
-   In practice, we evaluate treatment effects over a range of plausible values for $\lambda$.

This approach allows us to **bound the treatment effect** based on assumptions about the degree of omitted variable bias.

------------------------------------------------------------------------

The choice of $\lambda$ is crucial and inherently subjective, but some guidelines and empirical precedents help:

-   Small values of $\lambda$ (e.g., 0--1) represent modest levels of bias and suggest that selection on unobservables is not dominant.
-   Large values (e.g., $\lambda > 2$ or higher) imply that any omitted variables must be far more predictive of treatment than the observed controls, a strong and often implausible assumption in well-specified models.

This makes RCR particularly useful for *stress testing* causal estimates: "How bad would unobserved selection have to be to overturn my result?"

------------------------------------------------------------------------

This method has been applied to various marketing and digital strategy settings to assess the robustness of estimated effects. A few notable examples:

+------------------------+---------------------------------------------------------------------------+-------------------------------------+
| Study                  | Context                                                                   | Minimum $\lambda$ to nullify effect |
+========================+===========================================================================+=====================================+
| [@manchanda2015social] | **Social dollar effect**: impact of peer influence on purchasing behavior | 3.23                                |
+------------------------+---------------------------------------------------------------------------+-------------------------------------+
| [@chae2023paywall]     | **Paywall suspension**: effect on future subscription behavior            | 6.69                                |
+------------------------+---------------------------------------------------------------------------+-------------------------------------+
| [@sun2013ad]           | **Ad revenue-sharing**: impact on content popularity                      |                                     |
+------------------------+---------------------------------------------------------------------------+-------------------------------------+

These high $\lambda$ values imply that unobserved selection would need to be 3 to 7 times stronger than observable selection to eliminate the estimated treatment effect. In practical terms, this offers strong evidence that the effects are robust to omitted variable bias.

------------------------------------------------------------------------

We can estimate Relative Correlation Restrictions using the `rcrbounds` package in R, developed by @krauth2016bounding. This package estimates the bounds of the treatment effect across a range of $\lambda$ values and provides inference on whether the effect would remain statistically significant. Figure \@ref(fig:mm-rcrbounds-chickweight) plots the estimated treatment-effect bounds across $\lambda$ for the ChickWeight illustration.

```{r mm-rcrbounds-chickweight, fig.cap="Relative-correlation-restriction bounds for the treatment effect of Time on weight in the ChickWeight data, plotted across the sensitivity parameter lambda using the rcrbounds package."}
# Install if necessary:
# remotes::install_github("bvkrauth/rcr/r/rcrbounds")
# Declare the Python backend so reticulate resolves it per session.
reticulate::py_require("rcrbounds")
library(rcrbounds)

# Example using ChickWeight dataset
data("ChickWeight")

# Estimate treatment effect of Time on weight, controlling for Diet
rcr_res <- rcrbounds::rcr(
    formula = weight ~ Time | Diet, 
    data = ChickWeight, 
    # rc_range = c(0, 10) # Test lambda from 0 to 10
    rc_range = c(0, 1)
)

# Print summary
summary(rcr_res)

# Test whether the treatment effect is significantly different from 0
rcrbounds::effect_test(rcr_res, h0 = 0)

# Plot the bounds of the treatment effect
plot(rcr_res)
```

The plot shows how the estimated treatment effect varies as we allow for stronger selection on unobservables (i.e., increasing $\lambda$). If the effect remains consistently different from zero even at high $\lambda$, it provides graphical evidence of robustness.

------------------------------------------------------------------------

### 계수 안정성 경계 {#sec-coefficient-stability-bounds}

Robust regression analysis requires methods that account for potential **omitted variable bias**. A contribution to this literature is the coefficient-stability framework developed by @oster2019unobservable, which builds on the proportional selection model of @altonji2005selection. @oster2019unobservable derives conditions under which we can bound the bias due to unobservables using movements in the coefficient of interest and changes in model fit, as measured by $R^2$.

More recently, @masten2022effect have revealed that Oster's commonly used "delta" metric is insufficient to guarantee robustness of the **sign** of the treatment effect. They propose distinguishing between two types of breakdown points: one where the effect becomes zero (explain-away), and another where it changes sign.

Suppose the data‑generating process for outcome $Y$ is

$$
Y
=\beta X+\underbrace{W_1^{\!\top}\gamma_1}_{\text{observed}}
+\underbrace{W_2^{\!\top}\gamma_2}_{\text{unobserved}}
+\varepsilon ,
\quad
\operatorname{E}[\varepsilon|X,W_1,W_2]=0,
\tag{1}
$$

but the econometrician can only regress $Y$ on $X$ and the **observed** covariates $W_1$. By the omitted‑variable‑bias (OVB) formula,

$$
\widehat\beta_{\text{med}}-\beta
= \frac{\operatorname{Cov}(X,W_2^{\!\top}\gamma_2)}{\operatorname{Var}(X)} .
\tag{2}
$$

Because $W_2$ is hidden, nothing inside the sample alone fixes (2). The literature therefore introduces a *sensitivity parameter* that gauges how strongly $X$ is correlated with the missing variables.

@altonji2005selection postulated that the *selection on unobservables* is proportional to the *selection on observables*. @oster2019unobservable formalized the idea:

$$
\frac{\operatorname{Cov}(X,W_2^{\!\top}\gamma_2)}
     {\operatorname{Var}(W_2^{\!\top}\gamma_2)}
=\delta
\frac{\operatorname{Cov}(X,W_1^{\!\top}\gamma_1)}
     {\operatorname{Var}(W_1^{\!\top}\gamma_1)} .
\tag{3}
$$

where

-   $\delta$ is the **selection ratio**.
-   $\delta = 1$ says "unobservables are as correlated with $X$ as observables," yielding the canonical "Altonji‑Elder‑Taber benchmark."

@oster2019unobservable also assumes **coefficient alignment** ($\gamma_1=C\pi_1$) so we can link bias to the *movement* of $\widehat\beta$ and $R^2$ when controls are added. Under (1)--(3) the *long‑run* coefficient (with $W_2$ observed) satisfies

$$
\boxed{
\beta_{\text{long}}
=\beta_{\text{med}}
+(\beta_{\text{med}}-\beta_{\text{short}})
     \frac{R^2_{\text{long}}-R^2_{\text{med}}}
          {R^2_{\text{med}}-R^2_{\text{short}}}
}
\tag{4}
$$ where "short" means no controls, "med" means with $W_1$, and "long" is the unfeasible full model. Equation (4) is the **coefficient‑stability adjustment**. The two sensitivity objects are

-   **Bias‑adjusted coefficient**

    $$
    \beta^*(\delta,R^2_{\max})\equiv \beta_{\text{long}}
    \quad\text{after substituting }R^2_{\text{long}}=R^2_{\max},
    \delta\text{ into (4);} \tag{5}
    $$

-   **Breakdown point**

    $$
    \delta^{\text{EA}}
    =
    \inf\{\delta:\beta^*(\delta,R^2_{\max})=0\}.
    \tag{6}
    $$

When $\delta^{\text{EA}}>1$, unobservables must be *more* predictive of $X$ than observables to wipe out the estimate.

------------------------------------------------------------------------

@masten2022effect prove a stronger result: under Assumptions (1)--(3) the **sign‑change breakdown**

$$
\delta^{\text{SC}}
=\inf\{\delta:\operatorname{sign}\beta^*(\delta,R^2_{\max})
                \neq \operatorname{sign}\beta_{\text{med}}\}
\tag{7}
$$

is *always* ≤ 1 if $R^2_{\max}>R^2_{\text{med}}$ and $\beta_{\text{short}}\neq\beta_{\text{med}}$. Hence the traditional "$\delta = 1$ rule" can never guarantee that the *direction* of the effect is robust; it merely speaks to *magnitude*. Putting both $\delta^{EA}$ and $\delta^{SC}$ on the table is therefore critical.

------------------------------------------------------------------------

@oster2019unobservable recommends $R^2_{\max}=1.3R^2_{\text{med}}$ based on external evidence from randomized experiments, but researchers should vary this choice, especially if $R^2_{\text{med}}$ is modest. @cinelli2020making re‑express the problem in partial‑$R^2$ space and propose the **robustness value**, implemented in the `sensemakr` package. @imbens2003sensitivity supply complementary contour‑plot tools.

------------------------------------------------------------------------

```{r cs_robomit, message=FALSE, warning=FALSE}
library(tidyverse)
library(MatchIt)
library(robomit)


data("lalonde", package = "MatchIt")
lalonde <- lalonde %>% mutate(log_re78 = log(re78 + 1))

# create race dummies if needed
if (!("black" %in% names(lalonde))) {
  lalonde <- lalonde %>% mutate(
    black  = as.integer(race == "black"),
    hispan = as.integer(race == "hispanic")
  )
}

# nodegree naming patch
if (!("nodegr" %in% names(lalonde))) {
  if ("nodegree" %in% names(lalonde)) {
    lalonde <- lalonde %>% rename(nodegr = nodegree)
  }
}

# assure 0/1 indicators
lalonde <-
    lalonde %>% mutate(across(c(treat, black, hispan, married, nodegr),
                              as.numeric))

## analysis sample & R^2
vars <- c(
    "log_re78",
    "treat",
    "age",
    "educ",
    "black",
    "hispan",
    "married",
    "nodegr",
    "re74",
    "re75"
)
model_df <- lalonde %>% dplyr::select(all_of(vars)) %>% na.omit()

R2_med <- summary(
    lm(
        log_re78 ~ treat + age + educ + black + hispan +
            married + nodegr + re74 + re75,
        data = model_df
    )
)$r.squared
R2_max <- 1.3 * R2_med # Oster default upper bound

## β* (δ = 1) & δ* (β = 0) 
beta_tbl  <- o_beta(
    y = "log_re78",
    x = "treat",
    con = "age + educ + black + hispan + married + nodegr + re74 + re75",
    delta = 1,
    R2max = R2_max,
    type = "lm",
    data = model_df
)

delta_tbl <- o_delta(
    y = "log_re78",
    x = "treat",
    con = "age + educ + black + hispan + married + nodegr + re74 + re75",
    beta = 0,
    R2max = R2_max,
    type = "lm",
    data = model_df
)

beta_val  <- beta_tbl  %>% filter(Name == "beta*")  %>% pull(Value)
delta_val <- delta_tbl %>% filter(Name == "delta*") %>% pull(Value)

list(
  bias_adjusted_beta = beta_val,
  explain_away_delta = delta_val
)
```

```{r, message=FALSE, warning=FALSE}
# 1.  δ* as Rmax varies (explain‑away curve)
o_delta_rsq_viz(
  y   = "log_re78", x = "treat",
  con = "age + educ + black + hispan + married + nodegr + re74 + re75",
  beta = 0,                     # explain‑away target
  type = "lm",
  data = model_df
)

# 2.  bootstrap sampling distribution of δ*
o_delta_boot_viz(
  y   = "log_re78", x = "treat",
  con = "age + educ + black + hispan + married + nodegr + re74 + re75",
  beta   = 0,
  R2max  = R2_max,
  sim    = 100,                 # number of bootstrap draws
  obs    = nrow(model_df),      # draw full‑sample size each time
  rep    = TRUE,                # with replacement
  CI     = c(90, 95, 99),       # show three confidence bands
  type   = "lm",
  norm   = TRUE,                # overlay normal curve
  bin    = 120,                 # histogram bins
  data   = model_df
)

# 3.  bias‑adjusted β* over a grid of Rmax value
o_beta_rsq_viz(
  y   = "log_re78", x = "treat",
  con = "age + educ + black + hispan + married + nodegr + re74 + re75",
  delta = 1,                    # proportional selection benchmark
  type  = "lm",
  data  = model_df
)

```

1.  **Explain‑away curve**: plots $\delta^*$ against a range of $R^2_{\max}$.
2.  **Bootstrap histogram**: sampling distribution of $\delta^*$, with 90/95/99 % bands.
3.  $\beta^*$ vs. $R^2_{\max}$: how the bias‑adjusted coefficient moves as you tighten or loosen the assumed upper bound on $R^2$.
